\documentclass{article}
\usepackage[margin=1in]{geometry}
\usepackage[T1]{fontenc}
\usepackage[utf8]{inputenc}
\usepackage{microtype}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{mathtools}
\usepackage{hyperref}
\hypersetup{hidelinks}
\usepackage{etoolbox}
\usepackage{needspace}
\clubpenalty=10000
\widowpenalty=10000
\displaywidowpenalty=10000
\brokenpenalty=10000
\emergencystretch=2em
\usepackage[bottom]{footmisc}
\interfootnotelinepenalty=10000
\usepackage{fancyhdr}
\pagestyle{fancy}
\fancyhf{}
\lhead{Manifold Papers Collection}
\rhead{\thepage}
\allowdisplaybreaks
\numberwithin{equation}{section}
\setcounter{tocdepth}{1}
\newcommand{\R}{\mathbb{R}}
\newcommand{\T}{\mathbb{T}}
\newcommand{\dd}{\mathrm{d}}
\newcommand{\grad}{\nabla}
\pretocmd{\section}{\clearpage\Needspace{12\baselineskip}}{}{}
\pretocmd{\subsection}{\Needspace{8\baselineskip}}{}{}
\pretocmd{\subsubsection}{\Needspace{6\baselineskip}}{}{}
\pretocmd{\section*}{\clearpage\Needspace{12\baselineskip}}{}{}
\pretocmd{\subsection*}{\Needspace{8\baselineskip}}{}{}
\pretocmd{\subsubsection*}{\Needspace{6\baselineskip}}{}{}
\begin{document}

\title{Manifold Papers Collection}
\author{Joaquin Stürtz}
\date{January 2026}
\maketitle
\tableofcontents

\section{The Geodesic Flow Hypothesis: Unifying Memory and Computation via Differential Geometry}
\subsection*{Author}
Joaquin Stürtz
\subsection*{Date}
January 24, 2026
\subsection*{Abstract}
Contemporary deep learning architectures rely on discrete, layer-wise transformations that lack a coherent physical grounding, leading to significant challenges in modeling long-range dependencies and achieving energy-efficient scaling. We propose the Geodesic Flow Hypothesis, a foundational manifesto for artificial intelligence that reformulates neural computation as the continuous evolution of a state on a structured Riemannian manifold. We identify a fundamental ``Inertia-Accumulation Dualism'' and argue that intelligence requires the dynamic hybridization of Newtonian (conservative) and Aristotelian (dissipative) physical regimes. By treating reasoning as a geodesic flow governed by a learned metric and a state-dependent friction term---termed the Thermodynamic Clutch---we provide a framework for architectures that compute via dissipation and remember via conservation. The reference implementation in this repository (Manifold/GFN) realizes this by evolving position and velocity states, using low-rank Christoffel operators, learned dissipation gates, and geometric integrators.

\subsection{The Inertia-Accumulation Dualism}
We identify a central paradox in the physics of sequence modeling, arising from the conflict between the requirements for long-term storage and instantaneous update.

\subsubsection{The Newtonian Limit (Pure Inertia)}
In a purely conservative Hamiltonian system, information is stored in the momentum of the state particle. While this allows for infinite-horizon memory persistence (via Liouville's Theorem), it introduces extreme order sensitivity. In a second-order system $(\ddot{x} = F)$, early inputs exert disproportionate leverage over the final position compared to late inputs, making the learning of commutative operations (e.g., summation or parity) numerically unstable.

\subsubsection{The Aristotelian Limit (Pure Dissipation)}
In an over-damped system where velocity is proportional to force $(v \propto F)$, the state acts as a perfect first-order accumulator $(\dot{x} \approx F)$. This architecture excels at discrete computation but fails at long-range transmission: without an active force, the state remains frozen or decays, losing the ``inertial momentum'' required to carry context across temporal gaps.

\paragraph{Proposition}
Neither purely conservative nor purely dissipative dynamics are sufficient for general intelligence. A robust cognitive engine must be capable of Dynamic Regime Switching.

\subsection{Theoretical Framework: The Thermodynamic Clutch}
We propose that intelligence emerges from the dynamic modulation of the Dissipation Operator $\mu$. The unified equation of motion for a neural state $x$ is given by:
\[
\frac{d^2 x}{dt^2} + \Gamma_\theta(x, v)\cdot (v, v) = F_{ext}(t) - \mu_\theta(x, u)\cdot v
\]
Where $\Gamma_\theta(x, v)$ represents the geometry of the state space, $F_{ext}(t)$ the input force, and $\mu_\theta(x, u)$ a learnable, non-linear coupling factor that regulates energy dissipation. In Manifold, $\Gamma_\theta$ is parameterized by low-rank Christoffel operators and the dissipation term is produced by learned gates that depend on the current state (and optionally the input force). The integrator update uses the effective resistance $\Gamma_\theta(v,v) + \mu_\theta v$ and advances $(x, v)$ with a geometric step.

\subsubsection{The Two Regimes of Reasoning}
The model optimizes its parameters such that the system operates in two distinct physical states:
\begin{itemize}
\item High-Dissipation Regime (The Engaged Clutch): When $\mu \gg 1$, the system acts as a first-order accumulator, absorbing input forces into the spatial coordinate $x$. This is the regime of Active Computation.
\item Zero-Dissipation Regime (The Disengaged Clutch): When $\mu \approx 0$, the system acts as a second-order conservative flow, allowing the state to coast through the manifold via inertia. This is the regime of Information Persistence.
\end{itemize}
Cognitive performance is thus a function of the model's ability to ``engage the clutch'' only when relevant symbolic information is present, and ``disengage'' it to preserve that information over time. In practice, this appears as sparse, state-dependent spikes in $\mu_\theta$ rather than constant damping.

\subsection{Topology and Representation}
\subsubsection{Compact Manifolds and Topological Protection}
To prevent numerical explosion in second-order flows, the state space can be embedded in a compact topological variety. The repository implements an optional Hyper-Torus $T^n$ topology, which provides a bounded coordinate space $[0, 2\pi)^n$ for infinite-horizon coasting. In this geometry, logical states are protected by Winding Numbers---discrete topological invariants that are robust to numerical noise and perturbative drift. Euclidean topology remains available for tasks without periodic structure.

\subsubsection{Holographic Isomorphism}
We propose a Geometric Isomorphism between the latent state and the observable output. In the implementation this is optional: Manifold supports a holographic mode where the latent state is treated as the answer, and a standard mode where a lightweight readout projects the final position to logits. This preserves the interpretability advantages of holography while retaining practical flexibility for high-dimensional vocabularies.

\subsection{Conclusion}
The Geodesic Flow Hypothesis suggests that memory and computation are physically orthogonal processes. Memory is the byproduct of energy conservation (Hamiltonian physics), while computation is the byproduct of irreversibility and energy dissipation (Entropy/Thermodynamics). By aligning neural architectures with these fundamental physical laws, we enable the development of machines that do not merely ``process'' data, but physically resonate with the semantic structure of the world.

\subsection*{References}
[1] Arnold, V. I. (1989). Mathematical Methods of Classical Mechanics. Springer.\\
[2] Landauer, R. (1961). Irreversibility and Heat Generation in the Computing Process. IBM Journal of Research and Development.\\
[3] Bengio, Y., Simard, P., \& Frasconi, P. (1994). Learning long-term dependencies with gradient descent is difficult. IEEE Transactions on Neural Networks.\\
[4] Chen, R. T. Q., et al. (2018). Neural Ordinary Differential Equations. NeurIPS.\\
[5] Friston, K. (2010). The free-energy principle: a rough guide to the brain? Nature Reviews Neuroscience.\\
[6] Cover, T. M., \& Thomas, J. A. (2012). Elements of Information Theory. John Wiley \& Sons.\\
[7] Riemann, B. (1854). On the Hypotheses Which Lie at the Bases of Geometry.\\
[8] Hopfield, J. J. (1982). Neural networks and physical systems with emergent collective computational abilities. PNAS.

\section{The Hyper-Torus: Topology-Aligned Geometry for Constant-Memory Sequence Modeling}
\subsection*{Author}
Joaquin Stürtz
\subsection*{Date}
January 24, 2026
\subsection*{Abstract}
Cyclic reasoning tasks (parity, modular arithmetic, periodic phase tracking) are poorly expressed in Euclidean latent spaces. We present the Hyper-Torus, a topology-aligned manifold design that embeds the latent state into a product of circles $T^n = (S^1)^n$ with a physically meaningful Riemannian metric. Token processing becomes forced geodesic motion with symplectic discretization and periodic boundary conditions. The resulting architecture preserves information as phase and momentum, achieving constant-memory inference and long-horizon stability. We formalize the geometry, derive the geodesic equations (including Christoffel interactions and friction gates), describe the discretization via energy-preserving integrators, and detail training losses for periodic targets. Empirical evaluations on algorithmic tasks demonstrate robust extrapolation and reduced drift relative to non-geometric baselines.

\subsection{Motivation and Overview}
Modern sequence models often store an explicit history (e.g., KV caches), causing inference memory to grow with sequence length. Many algorithmic tasks, however, are naturally expressed as phase evolution (e.g., XOR parity as a half-rotation; counters as winding numbers). In such settings:
\begin{itemize}
\item Euclidean embeddings require continuous effort to remain on cyclic trajectories.
\item Periodicity is brittle near boundaries, leading to drift and aliasing.
\end{itemize}
The Hyper-Torus addresses these issues by aligning the manifold with the task: states are represented as phases on $S^1$, coupled into $T^n$, and evolved by physically structured updates. Symplectic discretization preserves qualitative dynamics; periodic boundary conditions remove edge discontinuities; friction gates implement controlled forgetting for regime switching.

\subsection{Geometric Preliminaries}
\subsubsection{The Torus}
We define the latent configuration space as a product of circles:
\[
\mathcal{M} \cong T^n = \underbrace{S^1 \times \cdots \times S^1}_{n\ \text{times}}
\]
with local coordinates $x = (\theta_1,\phi_1,\theta_2,\phi_2,\dots)$ in paired blocks. Each pair $(\theta,\phi)$ encodes a two-phase subsystem (inner/outer angles).

\subsubsection{Riemannian Metric}
Following a standard torus of revolution, we adopt a diagonal metric
\[
g(\theta,\phi) = \operatorname{diag}\big(r^2,\ (R + r\cos\theta)^2\big)
\]
and tile it across coordinates, yielding a block-diagonal $g(x)$ for $T^n$. Here $R$ is the major radius (global scale) and $r$ the minor radius (local scale). This geometry produces curvature terms that act as stabilizing geometric forces.

\subsubsection{Geodesic Equations with Forcing and Dissipation}
Let $x \in \mathcal{M}$ and velocity $v = \dot{x}$. The forced geodesic dynamics are
\[
\ddot{x}^k + \Gamma^k_{ij}(x)\,\dot{x}^i\dot{x}^j + \mu^k(x,u)\,\dot{x}^k = F^k(x,u)
\]
where $\Gamma$ are Christoffel symbols induced by $g$, $F$ is a token-driven force, and $\mu\ge 0$ is a dissipation coefficient (clutch) that gates between conservative memory and rapid rewriting. For the diagonal torus metric, representative non-zero symbol patterns for each $(\theta,\phi)$ block include
\[
\Gamma^{\theta}_{\phi\phi} \propto (R + r\cos\theta)\,\sin\theta / r,\quad
\Gamma^{\phi}_{\theta\phi} = \Gamma^{\phi}_{\phi\theta} \propto -\frac{r\sin\theta}{R + r\cos\theta}
\]
producing centrifugal and Coriolis-like effects that naturally regulate phase motion.

\subsection{Periodic Boundary Conditions and Distances}
\subsubsection{Boundary Wrapping}
Periodic coordinates are wrapped modulo $2\pi$ component-wise:
\[
x \leftarrow x \bmod 2\pi
\]
This removes discontinuities at chart boundaries and prevents unbounded drift.

\subsubsection{Toroidal Distance}
Prediction targets that live on $T^n$ use the shortest angular distance
\[
d_{\text{torus}}(x_1,x_2) = \min\!\big(|\Delta|,\ 2\pi - |\Delta|\big),\quad \Delta = x_1-x_2
\]
and losses such as $L_{\text{torus}} = \|d_{\text{torus}}(x_{\text{pred}},x_{\text{target}})\|_2^2$ or the bounded phase loss $L_{\text{phase}} = 1 - \cos(x_{\text{pred}} - x_{\text{target}})$ for smooth gradients near the wrap.

\subsection{Token-to-Force Mapping and Multi-Head Geometry}
Tokens are embedded into per-step forces $F_\theta(u_t)$ that act on $v$ and, via $\Gamma(x,v)$, deform trajectories. The latent state is split into $H$ heads. Each head owns a curvature module and integrator step; heads are mixed back linearly. This parallel decomposition yields:
\begin{itemize}
\item Independent sub-geometries for specialized reasoning channels.
\item Learned per-head time scales to adapt resolution to local complexity.
\end{itemize}

\subsection{Friction Gates and Reactive Curvature}
\subsubsection{Thermodynamic Gating (The Clutch)}
We parameterize dissipation as $\mu(x,u) = \sigma(W_{\text{state}}\,[\sin x,\cos x] + W_{\text{input}}\,u)$, where $[\sin x,\cos x]$ provides periodic features. Small $\mu$ keeps memory conservative; large $\mu$ rapidly damps momentum to overwrite state.

\subsubsection{Reactive Curvature}
Curvature can be modulated by kinetic energy $K=\tfrac12\|v\|^2$ via
\[
\Gamma_{\text{eff}}(x,v) = \Gamma_{\text{base}}(x) \cdot \big(1 + \alpha\,\tanh K\big)
\]
increasing geometric resistance in high-energy regimes and improving stability through difficult transitions.

\subsubsection{Singularity Potentials (Optional)}
Logical bottlenecks may be modeled as localized potentials that strengthen curvature near thresholds, imitating event horizons that stabilize discrete flips.

\subsection{Discretization: Energy-Preserving Integrators}
We discretize the dynamics with structure-preserving schemes:
\begin{itemize}
\item Symplectic/Verlet and Leapfrog (kick-drift-kick).
\item Higher-order symplectic compositions (Yoshida, Forest-Ruth, Omelyan) for smoother regimes.
\item Heun/RK2 as a robust non-symplectic baseline.
\end{itemize}
A typical Leapfrog step for $(x,v)$ with torus boundary is
\[
v_{\tfrac12} = v + \tfrac12\,dt\,a(x,v,u),\quad
x' = \operatorname{wrap}\big(x + dt\,(v_{\tfrac12})\big),\quad
v' = v_{\tfrac12} + \tfrac12\,dt\,a(x',v_{\tfrac12},u)
\]
where $a(x,v,u) = F(x,u) - \Gamma(x,v) - \mu(x,u)\,v$ and $\operatorname{wrap}(\cdot)$ applies the $2\pi$ periodic boundary. This preserves qualitative phase-space structure and prevents long-horizon drift.

\subsection{Parallelization (Optional)}
For training throughput, first-order affine approximations of the recurrence
\[
v_t = A_t\,v_{t-1} + b_t,\qquad
x_t = x_{t-1} + dt_t\,v_t
\]
can be evaluated via associative parallel scans on GPUs. This reduces serial depth while maintaining functional agreement with sequential updates.

\subsection{Training Objectives for Periodic Targets}
We combine task losses with physics-informed regularization:
\begin{itemize}
\item Cross-entropy or toroidal/circular distance losses for outputs on $T^n$.
\item Hamiltonian loss $L_H$ to penalize spurious energy creation:
\[
L_H = \lambda_H\,\mathbb{E}\big[\,|E_{t+1} - E_t|\,\big],\quad
E_t = \tfrac12\,v_t^\top g(x_t)\,v_t
\]
\item Geodesic curvature regularization to temper excessive curvature excursions.
\item Optional soft-symmetry (Noether) terms to align isomeric heads.
\end{itemize}

\subsection{Complexity and Memory}
Hyper-Torus maintains a fixed-size state $(x_t,v_t)$ independent of sequence length $L$:
\[
\text{Memory} = \mathcal{O}(1),\qquad
\text{Compute} \approx \mathcal{O}(L\,d^2)\ \text{(dense mixing)}
\]
Symplectic discretization reduces gradient pathologies; periodic wrapping eliminates boundary artifacts, supporting stable infinite-horizon reasoning.

\subsection{Empirical Behavior}
On cyclic algorithmic tasks (e.g., cumulative parity), Hyper-Torus exhibits:
\begin{itemize}
\item Stable phase tracking visualized as limit cycles on periodic coordinates.
\item Strong extrapolation far beyond training sequence lengths.
\item Reduced drift versus non-geometric baselines through boundary-aware updates and energy-preserving integration.
\end{itemize}

\subsection{Discussion and Limitations}
While torus topology aligns well with cyclic reasoning, non-smooth dynamics (sharp logical transitions) can challenge high-order explicit schemes; lower-order symplectic or robust RK2 steps often perform best. Learned friction and reactive curvature provide practical stabilization but introduce task-dependent hyperparameters. Formal convergence proofs for learned curvature remain open research.

\subsection{Conclusion}
Hyper-Torus reframes sequence modeling as phase-space evolution on $T^n$. By matching topology to task and respecting geometry in discretization, the architecture preserves information as momentum and winding rather than as explicit memory buffers. This yields constant-memory inference, robust long-horizon stability, and a principled path to physically grounded neural reasoning.

\subsection*{References}
[1] Riemann, B. (1854). Über die Hypothesen, welche der Geometrie zu Grunde liegen.\\
[2] Arnold, V. I. (1989). Mathematical Methods of Classical Mechanics.\\
[3] Hairer, E., Lubich, C., \& Wanner, G. (2006). Geometric Numerical Integration.\\
[4] Chen, R. T. Q., et al. (2018). Neural Ordinary Differential Equations.\\
[5] Bronstein, M. M., et al. (2021). Geometric Deep Learning.\\
[6] MacKay, D. J. C. (2003). Information Theory, Inference, and Learning Algorithms.\\
[7] Yoshida, H. (1990). Construction of higher order symplectic integrators.\\
[8] Omelyan, I. P., et al. (2002). Symplectic algorithms for molecular dynamics equations.\\
[9] Dinh, L., et al. (2014). NICE: Non-linear Independent Components Estimation.\\
[10] Rezende, D. J., \& Mohamed, S. (2015). Variational Inference with Normalizing Flows.

\section{Reactive Geometry: Energy-Modulated Curvature for Stable Neural Geodesic Flow}
\subsection*{Author}
Joaquin Stürtz
\subsection*{Date}
January 24, 2026
\subsection*{Abstract}
We introduce Reactive Geometry, a principled extension of Riemannian modeling in which the latent manifold stiffens in response to the instantaneous kinetic energy of the state. The metric acts as a self-regulating substrate: high energy (semantic uncertainty) increases curvature and friction to brake chaotic motion; low energy (semantic certainty) allows inertial geodesic reasoning. Concretely, reactive curvature appears as a bounded multiplicative scaling of Christoffel interactions computed from a low-rank geometric operator, combined with state- and input-dependent dissipation gates and optional singularity potentials that stabilize discrete logical transitions. We present the mathematical formulation, derive the effective acceleration, detail structure-preserving discretizations, and outline training objectives aligned with periodic targets and energy conservation.

\subsection{Introduction}
Riemannian manifolds provide structure for latent computation, yet fixed geometries cannot capture the subjective effort of reasoning. In Reactive Geometry, the manifold responds to the state co-vector’s energy, producing a closed-loop control: energy $\to$ curvature $\to$ braking $\to$ reduced energy. This establishes a physically grounded path to uncertainty handling and long-horizon stability in neural geodesic flow.

\subsection{Formalism}
\subsubsection{Low-Rank Curvature Operator}
Let $x\in\mathbb{R}^d$ be coordinates and $v=\dot{x}$ the velocity. The Christoffel interaction is parameterized by a low-rank, symmetric operator:
\[
\Gamma^k_{\!ij}(x) \approx \sum_{r=1}^{R}\,\lambda_{kr}\,U_{ir}\,U_{jr}
\]
with diagonal saturation to bound extreme values. In practice, the projection $p=vU$ is used to scale curvature, ensuring numerical robustness by attenuating responses with a factor like $1/(1+\|p\|)$, and applying a smooth saturation (e.g., $\tanh$) to clip large forces.

\subsubsection{Reactive Curvature (Plasticity)}
The effective connection multiplies base curvature by a bounded function of kinetic energy:
\[
\Gamma_{\text{eff}}(x,v) = \Gamma_{\text{base}}(x)\,\big(1+\alpha\,\tanh K\big),
\qquad K=\tfrac12\,\|v\|^2
\]
where $\alpha\ge 0$ is a plasticity coefficient. As $K$ increases, curvature stiffens and turns motion more strongly, acting as a self-braking mechanism that reduces oscillations and runaway drift.

\subsubsection{Dissipation Gates (The Clutch)}
Reactive Geometry uses a conformal symplectic dissipation term to regulate state rewriting:
\[
\mu(x,u) = \sigma\!\big(W_{\text{state}}\,[\sin x,\cos x] + W_{\text{input}}\,u\big)\cdot \mu_{max}
\]
with periodic features $[\sin x,\cos x]$ for toroidal topologies and an input-dependent component coupled to the token force $u$. Small $\mu$ keeps memory conservative; large $\mu$ damps momentum to allow rapid overwriting.

\subsubsection{Singularity Potentials (Optional)}
Discrete logical flips can be stabilized by localized potentials:
\[
S(x) = \sigma\!\big(V(x)\big),\qquad
\Gamma_{\text{eff}} \leftarrow \Gamma_{\text{eff}}\cdot\!\big(1+(S(x)-S_0)\,(\beta-1)\big)
\]
where $V$ is a learned position-gate, $S_0$ a threshold, and $\beta>1$ the singularity strength. The effect is a soft, differentiable intensification of curvature near high-certainty regions, forming event horizons that trap the state after decisive transitions.

\subsubsection{Effective Acceleration}
The latent dynamics combine forcing, curvature, and dissipation:
\[
a(x,v,u) = F(x,u) - \Gamma_{\text{eff}}(x,v) - \mu(x,u)\,v
\]
where $F$ is a token-to-force mapping. This form yields a conservative memory mode $(\mu\approx 0)$ and a dissipative rewrite mode $(\mu\gg 0)$, both modulated by reactive curvature from $K$.

\subsection{Geometry–Topology Interplay}
Reactive Geometry is compatible with non-Euclidean topologies. In toroidal settings:
\begin{itemize}
\item Coordinates are periodic $(x\bmod 2\pi)$, preventing edge discontinuities.
\item Periodic features $[\sin x,\cos x]$ feed friction and potential gates.
\item The diagonal torus metric blocks (inner/outer angles) produce centrifugal/Coriolis-like terms that naturally regulate phase motion.
\end{itemize}

\subsection{Discretization and Stability}
\subsubsection{Structure-Preserving Integrators}
Reactive dynamics are discretized via energy-preserving schemes:
\begin{itemize}
\item Leapfrog/Verlet (kick-drift-kick) for robust qualitative stability.
\item Higher-order symplectic compositions (Yoshida, Forest-Ruth, Omelyan) in smooth regimes.
\item Heun/RK2 as a non-symplectic but stable baseline.
\end{itemize}
A typical Leapfrog step with periodic wrapping is:
\[
v_{\tfrac12}=v+\tfrac12\,dt\,a(x,v,u),\quad
x'=\operatorname{wrap}\!\big(x+dt\,v_{\tfrac12}\big),\quad
v'=v_{\tfrac12}+\tfrac12\,dt\,a(x',v_{\tfrac12},u)
\]

\subsubsection{Saturation and Clamps}
To avoid numerical explosion under non-smooth forces, curvature outputs are softly saturated (e.g., via $\tanh$) and clamped within bounded ranges. Dissipation coefficients use sigmoid activation and a fixed scale to keep updates within safe limits. These measures, together with reactive modulation, reduce oscillations and help maintain energy budgets.

\subsubsection{Time-Scale Adaptation}
Reactive Geometry can be paired with learned per-head time scales and optional dynamic time gating, reducing $dt$ in complex regions and increasing it in near-flat geometry, improving stability without extra memory.

\subsection{Implementation Overview}
\begin{itemize}
\item Curvature operator: low-rank symmetric parameterization with projection-based scaling and smooth saturation.
\item Reactive curvature: multiplicative plasticity using a bounded function of kinetic energy.
\item Friction gates: state-dependent via periodic features and input-dependent via the force; both combined through sigmoid activation.
\item Singularity potentials: differentiable intensification of curvature near high-certainty regions using a learned position gate and soft thresholding.
\item Topology: periodic wrapping for toroidal coordinates; periodic features in gates and potentials; diagonal torus metric blocks for stabilizing interactions.
\item Execution paths: reliable Python loops for correctness; optional fused GPU kernels; optional scan-based parallelization for training throughput, preserving functional behavior with bounded numerical differences.
\end{itemize}

\subsection{Training Objectives}
Reactive Geometry integrates task losses with physics-informed regularization:
\begin{itemize}
\item Task loss: cross-entropy for discrete prediction, or periodic targets using toroidal or bounded phase losses.
\item Hamiltonian penalty: discourages spurious energy creation by penalizing $|E_{t+1}-E_t|$, with energy computed under the metric blocks $E_t=\tfrac12\,v_t^\top g(x_t)\,v_t$ when available.
\item Geodesic regularization: tempers curvature excursions via a mean-squared curvature term.
\item Symmetry regularization (optional): encourages consistent geometric responses across isomeric heads.
\end{itemize}

\subsection{Empirical Behavior}
Reactive Geometry consistently reduces drift by braking high-energy trajectories and stabilizing discrete transitions with optional singularity potentials. On cyclic algorithmic tasks, it supports long-horizon phase tracking and robust extrapolation, especially in combination with toroidal topologies and structure-preserving integrators. In non-smooth regimes, lower-order symplectic or RK2 steps often outperform high-order explicit schemes.

\subsection{Discussion and Limitations}
Reactive manifolds relate to Finslerian extensions by allowing velocity-dependent geometry, yet remain practical via simple bounded scaling of curvature and gates. Singularities provide useful stabilization for discrete logic but must be softly gated to preserve differentiability. CUDA fused paths and parallel scans yield throughput gains, though strict floating-point parity with sequential updates is not guaranteed. Formal convergence proofs for learned curvature and reactive flows are an open direction.

\subsection{Conclusion}
Reactive Geometry equips neural geodesic flow with an internal feedback loop: energy modulates curvature and friction, which in turn regulates energy. This yields stable long-horizon behavior, principled uncertainty handling, and improved alignment between geometry and computation. Combined with topology-aware modeling and structure-preserving discretization, it forms a foundation for constant-memory, physically grounded sequence reasoning.

\subsection*{References}
[1] Friston, K. (2010). The free-energy principle: a unified brain theory?\\
[2] Amari, S. (2016). Information Geometry and Its Applications.\\
[3] Einstein, A. (1916). The Foundation of the General Theory of Relativity.\\
[4] Bao, D., Chern, S. S., \& Shen, Z. (2000). An Introduction to Riemann-Finsler Geometry.\\
[5] Sitzmann, V., et al. (2020). Implicit Neural Representations with Periodic Activation Functions.\\
[6] Nicolis, G., \& Prigogine, I. (1977). Self-Organization in Non-Equilibrium Systems.\\
[7] Bejan, A. (2000). Shape and Structure, from Engineering to Nature.

\section{Thermodynamic Gating: Dissipation-Controlled Memory in Geodesic Flow Networks}
\subsection*{Author}
Joaquin Stürtz
\subsection*{Date}
January 24, 2026
\subsection*{Abstract}
Hamiltonian latent dynamics preserve phase-space volume, enabling long-horizon persistence of information. Yet strict conservation impedes contextual switching: a purely conservative system cannot relax into new semantic attractors without oscillation. We introduce Thermodynamic Gating, a mechanism that couples Hamiltonian geodesic flow to a learned dissipative field and a dynamic time gate. A state- and input-dependent friction coefficient $\mu(x,u)$ implements selective irreversibility (the ``clutch''), while a curvature-informed gate scales the effective time step $\Delta t_{\text{eff}} = g(x)\Delta t$. The result is a conformal-symplectic update that supports both persistent memory (coasting) and decisive rewriting (damping), verified in the reference implementation by an implicit friction Leapfrog scheme and toroidal topology features for periodic computation.

\subsection{Motivation: Conservation, Updating, and Selective Irreversibility}
\begin{itemize}
\item Conservation preserves gradients and memory but resists settling; naive damping destroys long-horizon stability.
\item Intelligent behavior requires switching between an Isolated Regime (conservative flow) and an Open Regime (entropy-producing update).
\item Thermodynamic Gating provides a physical basis for the forget gate: dissipation is produced only where and when information must be rewritten.
\end{itemize}

\subsection{Continuous-Time Model}
We augment Hamiltonian dynamics with a dissipative co-vector proportional to velocity:
\[
\dot{x} = v, \quad \dot{v} = F(x,u) - \Gamma(x,v) - \mu(x,u)\odot v
\]
where $F$ is an external force embedding (token-driven), $\Gamma$ is the Christoffel-like curvature interaction, and $\mu(x,u) \ge 0$ is a learned dissipation coefficient. Two regimes arise:
\begin{itemize}
\item Memory mode $(\mu \approx 0)$: near-Hamiltonian coasting (information persistence).
\item Update mode $(\mu \gg 0)$: rapid damping (information rewriting at transitions).
\end{itemize}
Additionally, a learned gate $g(x)\in(0,1]$ scales the effective step size, shrinking time in hard regions (high curvature) and expanding it in flat regions (skip-like behavior).

\subsection{Discrete-Time: Conformal Symplectic Leapfrog with Implicit Friction}
Let $\Delta t$ be the base step and $g(x)\in(0,1]$ the dynamic scale. Define $\Delta t_{\text{eff}} = g(x)\Delta t$ and $h = \tfrac{1}{2}\Delta t_{\text{eff}}$. A single Leapfrog step with implicit friction is:
\begin{enumerate}
\item Kick (half step, implicit damping):
\[
v_{n+\tfrac{1}{2}} = \frac{v_n + h\,[F(x_n,u_n) - \Gamma(x_n,v_n)]}{1 + h\,\mu(x_n,u_n)}
\]
\item Drift (full step position):
\[
x_{n+1} = \operatorname{wrap}\!\left(x_n + \Delta t_{\text{eff}}\; v_{n+\tfrac{1}{2}}\right)
\]
\item Kick (half step at new position):
\[
v_{n+1} = \frac{v_{n+\tfrac{1}{2}} + h\,[F(x_{n+1},u_n) - \Gamma(x_{n+1},v_{n+\tfrac{1}{2}})]}{1 + h\,\mu(x_{n+1},u_n)}
\]
\end{enumerate}
Here wrap$(\cdot)$ enforces topology (e.g., periodic torus). This implicit form is stable under large $\mu$ and matches the fused CUDA/Python reference update. In dimensions, $\mu$ may be per-coordinate; the gate is bounded by a sigmoid scale (maximum friction).

\subsection{Implementation Summary (Architecture-Level)}
\begin{itemize}
\item Friction Gate (Thermodynamic Clutch): A linear mapping over state features (and optionally input force) produces $\mu(x,u)=\alpha\cdot\sigma(\cdot)$ with $\alpha>0$. On compact manifolds (torus), features are periodic $[\sin(x), \cos(x)]$.
\item Curvature Gate (Dynamic Time): A small MLP over $x$ yields $g(x)\in(0,1]$, scaling $\Delta t$ per head. On torus, inputs use $[\sin(x), \cos(x)]$ to respect periodicity.
\item Curvature Interaction: A low-rank Christoffel operator computes $\Gamma(x,v)$ with symmetric structure to approximate torsion-free geometry; outputs are softly clamped for numerical safety.
\item Integrators: Leapfrog/Verlet/Yoshida/Forest–Ruth implement the geodesic step; Leapfrog uses the implicit friction update above and respects topology via periodic wrapping.
\item Multi-Head Manifolds: The latent state is split across heads; each head applies its own geometry, gate, and integrator, then results are mixed back.
\end{itemize}

\subsection{Topology and Periodic Features}
\begin{itemize}
\item Toroidal topology $T^n$ bounds coordinates and models cyclic computation naturally (parity, phase, modular arithmetic).
\item Periodic features $[\sin(x), \cos(x)]$ are used both in friction gating and dynamic time gating to preserve continuity across $2\pi$ boundaries.
\item Position updates apply wrap$(\cdot)$ to maintain coordinates on the manifold; training losses include toroidal distance terms to avoid boundary artifacts.
\end{itemize}

\subsection{Training and Regularization}
\begin{itemize}
\item Task losses (e.g., cross-entropy or toroidal distance) drive semantic objectives.
\item Hamiltonian stability terms penalize spurious energy creation during coasting segments.
\item Geodesic regularization encourages curvature smoothness and reduces instability under strong gates.
\item Velocity saturation (tanh) limits runaway speeds while keeping gradients well-behaved.
\end{itemize}

\subsection{Information-Theoretic Perspective}
Thermodynamic Gating realizes Landauer’s Principle: erasure requires dissipating energy. In practice, trained models exhibit:
\begin{itemize}
\item Low dissipation during long stationary contexts (memory conservation).
\item Sharp, localized spikes of $\mu$ at semantic transitions (entropy production where rewriting is required).
\item Dynamic time contraction near complex geometric regions to avoid aliasing and numerical instability.
\end{itemize}

\subsection{Practical Notes}
\begin{itemize}
\item Maximum friction scale is bounded to maintain stability under implicit updates.
\item Gates are per-head and per-batch, enabling heterogeneous regimes across manifold subspaces.
\item CUDA fused paths accelerate integration and gating; Python fallbacks preserve correctness where fused kernels are unavailable.
\item Toroidal wrapping is applied after position updates; gating features remain periodic to avoid discontinuities.
\end{itemize}

\subsection{Conclusion}
Thermodynamic Gating unifies conservative memory and decisive updating within a single geometric computation layer. By combining a conformal-symplectic integrator with learned friction and dynamic time gating, Geodesic Flow Networks achieve stability, persistence, and controllability on compact manifolds—enabling symbolic trajectory formation and robust reasoning over long horizons.

\subsection*{References}
[1] Prigogine, I. (1955). Introduction to Thermodynamics of Irreversible Processes. Thomas.\\
[2] Landauer, R. (1961). Irreversibility and Heat Generation in the Computing Process. IBM Journal of Research and Development.\\
[3] Greydanus, S., et al. (2019). Hamiltonian Neural Networks. NeurIPS.\\
[4] Hochreiter, S., \& Schmidhuber, J. (1997). Long Short-Term Memory. Neural Computation.\\
[5] Ottinger, H. C. (2005). Beyond Equilibrium Thermodynamics. Wiley-Interscience.\\
[6] Cranmer, M., et al. (2020). Lagrangian Neural Networks. ICLR.\\
[7] Schlögl, F. (1971). Thermodynamic stability of non-equilibrium states. Zeitschrift für Physik.

\section{Symplectic Attention: Constant-Memory Sequence Modeling via Geodesic Flow}
\subsection*{Author}
Joaquin Stürtz
\subsection*{Date}
January 24, 2026
\subsection*{Abstract}
Attention computes non-local interactions over a growing memory of keys and values, incurring quadratic time and linear memory with sequence length. We claim that this explicit global lookup is not necessary if the model’s recurrence obeys a physically grounded dynamical law. Symplectic Attention replaces pairwise token interactions with local geodesic flow on a learned manifold: the model carries its causal history in a compact phase state $(x_t, v_t)$, and ``attends'' by evolving along learned curvature with structure-preserving integrators. A learned friction gate provides selective irreversibility for state rewriting, while a curvature-informed time gate shrinks steps in complex regions. The result is constant-memory inference with linear time, verified by a reference implementation based on multi-head manifolds, toroidal topology, and symplectic integration.

\subsection{Motivation: From Global Lookup to Local Flow}
\subsubsection{The cost of Action at a Distance}
Transformers maintain a full history of Key–Value pairs $(K, V)$ and, to produce the next output, query this entire memory:
\[
\operatorname{Attention}(Q, K, V) = \operatorname{softmax}\!\left(\frac{QK^\top}{\sqrt{d}}\right)V
\]
This non-local operation scales quadratically in sequence length. It assumes relevant information must be explicitly retrieved from a static cache rather than propagated causally.

\subsubsection{The Physical Alternative}
In physics, a particle does not query distant points; it moves under local fields. The ``memory'' of structure is encoded in geometry (metric and connection), and the trajectory reveals which past impulses matter. We reformulate sequence modeling so that ``attention'' arises from geodesic deviation: the state flows along learned curvature that locally encodes how past tokens shape present motion.

\subsection{Continuous-Time Formalism}
We augment Hamiltonian dynamics with learned curvature and selective dissipation:
\[
\dot{x} = v, \quad \dot{v} = F_\theta(u_t) - \Gamma_\theta(x, v) - \mu_\theta(x, u_t)\odot v
\]
Here, $F_\theta$ maps tokens to force impulses, $\Gamma_\theta$ is a learned Christoffel-like interaction that encodes geometric coupling, and $\mu_\theta\ge 0$ is a friction clutch that switches between conservative memory and dissipative updating.

\subsubsection{Symplectic Attention as Geodesic Flow}
Instead of $\mathcal{O}(N^2)$ pairwise weighting, past tokens influence the present through the shape of the trajectory. The instantaneous phase state $(x_t, v_t)$ encodes causal history as momentum and winding on a compact manifold. Attending is realized by local curvature steering; dissipation opens the clutch to overwrite state during transitions.

\subsubsection{Multi-Head Manifolds}
Analogous to multi-head attention, we factor the latent into $H$ heads, each with its own geometry and gates. Heads evolve in parallel, then are mixed back into the shared state. This yields diverse relational channels without constructing global attention matrices.

\subsubsection{Dynamic Time and Selective Irreversibility}
A small gate $g(x)\in(0,1]$ scales the effective step size in hard regions (high curvature), while $\mu(x,u)$ controls when to dissipate momentum and rewrite the state. Together they implement stable, controllable computation over long horizons.

\subsubsection{Symplectic Stability}
For $\mu=0$, the update is symplectic; phase-space volume is preserved and gradients neither vanish nor explode. Forgetting arises physically through $\mu>0$, not numerically through unstable mappings.

\subsection{Discretization: Symplectic Attention Step}
Let $\Delta t$ be the base step, $g(x)\in(0,1]$ the time gate, $\Delta t_{\text{eff}} = g(x)\Delta t$, and $h=\tfrac{1}{2}\Delta t_{\text{eff}}$. A single head applies a Leapfrog (kick–drift–kick) step with implicit friction:
\begin{enumerate}
\item Kick (half-step, implicit damping):
\[
v_{n+\tfrac{1}{2}} = \frac{v_n + h\,[F_\theta(u_n) - \Gamma_\theta(x_n, v_n)]}{1 + h\,\mu_\theta(x_n, u_n)}
\]
\item Drift (full-step position with topology):
\[
x_{n+1} = \operatorname{wrap}\!\left(x_n + \Delta t_{\text{eff}}\, v_{n+\tfrac{1}{2}}\right)
\]
\item Kick (half-step at new position):
\[
v_{n+1} = \frac{v_{n+\tfrac{1}{2}} + h\,[F_\theta(u_n) - \Gamma_\theta(x_{n+1}, v_{n+\tfrac{1}{2}})]}{1 + h\,\mu_\theta(x_{n+1}, u_n)}
\]
\end{enumerate}
Heads evolve in parallel; their states are concatenated and linearly mixed back into $(x, v)$. The scheme generalizes to other symplectic integrators (Yoshida, Forest–Ruth, Omelyan) and high-order explicit schemes (Heun/RK4/RK45) when conservation needs trade off against local accuracy.

\subsubsection{The Infinite Context Limit}
Liouville’s Theorem implies that information about initial conditions is preserved in ideal symplectic systems; causal history becomes phase windings on compact manifolds. The challenge shifts from storage (global cache) to resolution (readout decoding of state).

\subsubsection{Complexity}
\begin{itemize}
\item Time: $\mathcal{O}(N)$ por secuencia (recurrencia constante por token).
\item Memory: $\mathcal{O}(1)$ en inferencia (estado de tamaño fijo).
\item Parallelization: loop-friendly over tokens; sequence scans can reduce training wall-time on accelerators.
\end{itemize}

\subsection{Architecture Mapping}
\subsubsection{Heads and Geometry}
Each head maintains a sub-state and a geometry module (learned low-rank curvature or specialized manifold). Geometries compute $\Gamma_\theta(x, v)$ and expose gates for $\mu(x,u)$; on compact topologies, gates use periodic features $[\sin(x), \cos(x)]$.

\subsubsection{Dynamic Time Gate}
A small MLP over state yields $g(x)\in(0,1]$ per head, scaling the effective step. This contracts time in high-curvature regions while allowing coasting elsewhere.

\subsubsection{Mixing and Readout}
After integration, heads are concatenated and linearly mixed back into the global $(x, v)$. Readout maps the latent to outputs via linear or implicit heads; in holographic mode, the latent state itself represents the target (e.g., phases), enabling topology-aligned supervision.

\subsubsection{Topology}
Toroidal topology bounds coordinates and encodes cyclic computation robustly. Position updates apply periodic wrap; gates consume periodic features to remain continuous at boundaries. Losses include toroidal or circular distance to avoid edge artifacts.

\subsubsection{Training}
In addition to task loss (e.g., cross-entropy or toroidal distance), physics-informed terms stabilize trajectories:
\begin{itemize}
\item Hamiltonian stability loss: penalizes spurious energy creation during coasting.
\item Geodesic regularization: smooths curvature excursions.
\item Velocity saturation: prevents runaway speeds while keeping gradients bounded.
\end{itemize}

\subsection{Empirical Notes}
On algorithmic tasks with long-range structure (parity, modular arithmetic, periodic phase tracking), models with compact topology and symplectic integration maintain stable trajectories and generalize to longer sequences. Dissipation spikes align with semantic transitions, while time gates shrink steps in geometrically complex regions, enabling precise updates without global attention matrices.

\subsection{Discussion and Conclusion}
Symplectic Attention returns to local causality: information moves through geometry, not global lookups. By carrying causal history in a compact phase state and evolving it with structure-preserving updates, we achieve constant-memory inference and robust long-horizon behavior. Selective irreversibility allows decisive state changes, while periodic topology aligns representation with cyclic tasks. This paradigm grounds sequence modeling in physics, offering a scalable alternative to quadratic attention.

\subsection*{References}
[1] Vaswani, A., et al. (2017). Attention Is All You Need. NeurIPS.\\
[2] Arnold, V. I. (1989). Mathematical Methods of Classical Mechanics. Springer-Verlag.\\
[3] Katharopoulos, A., et al. (2020). Transformers are RNNs: Fast Autoregressive Transformers with Linear Attention. ICML.\\
[4] Chen, R. T. Q., et al. (2018). Neural Ordinary Differential Equations. NeurIPS.\\
[5] Gu, A., \& Dao, T. (2023). Mamba: Linear-Time Sequence Modeling with Selective State Spaces. arXiv.\\
[6] Noether, E. (1918). Invariant Variation Problems. Nachrichten von der Gesellschaft der Wissenschaften zu Göttingen.\\
[7] Misner, C. W., Thorne, K. S., \& Wheeler, J. A. (1973). Gravitation. W. H. Freeman.

\section{Holographic Latent Space: Zero-Shot Readout via Intrinsic Geometric Alignment}
\subsection*{Author}
Joaquin Stürtz
\subsection*{Date}
January 24, 2026
\subsection*{Abstract}
Contemporary neural networks are ``Black Boxes'' primarily because their internal representations are separated from their outputs by opaque projection layers. We introduce Holographic Readout, a training mode that encourages the latent state to align directly with the target geometry. In the reference implementation this is optional: the model can use a lightweight readout head or a holographic mode where the latent position is treated as the answer. This compels the network to perform Intrinsic Geometric Alignment: the thought of a concept must maintain the same topological structure as the concept itself. On cyclic tasks, this is naturally represented as angular rotations (e.g., $\theta \in \{0, \pi\}$) in a toroidal coordinate system, improving interpretability by design.

\subsection{Introduction}
\subsubsection{The Translation Gap}
In a standard Transformer, the vector for ``dog'' inside the model looks nothing like a dog. It is a high-dimensional hash that only becomes meaningful when multiplied by a massive Unembedding Matrix. This separation creates two problems:
\begin{itemize}
\item Interpretability: We cannot inspect the thought process without decoding it first.
\item Robustness: The core reasoning engine can be hallucinating, but the projection layer might mask the error (or vice-versa).
\end{itemize}

\subsubsection{The Holographic Hypothesis}
We propose that a truly intelligent system should not need an interpreter. Its internal state should be isomorphic to the external reality it models.

If the task is to track a cyclic variable (0, 1, 0, 1...), the internal state should actually rotate. If the task is hierarchical, the state should branch.

We can enforce this by setting the Readout Layer to the Identity mapping:
\[
y_{pred} = x_{latent}
\]
This forces the entire deep network to act as a Differentiable Simulator rather than a statistical approximate function.

\subsection{Geometric Alignment Theory}
\subsubsection{The Alignment Loss}
Let the target space be a manifold $\mathcal{Y}$ (e.g., the circle $S^1$). Let the latent space be $\mathcal{X}$. Standard training minimizes $d(f(x), y)$. Holographic training minimizes $d(x, y)$ directly.

This creates a powerful inductive bias: The laws of physics inside the model ($F=ma$) must mimic the logical rules of the task.

\subsubsection{Case Study: The Parity Torus}
In the Parity task (XOR sum), the target is binary $\{0, 1\}$. We map this to the Torus $S^1$:
\begin{itemize}
\item $0 \to 0$ radians
\item $1 \to \pi$ radians
\end{itemize}
If we enforce Holographic Readout, the network cannot simply classify the input. It must move its latent particle from $0$ to $\pi$. Input ``1'' acts as a force $F$. The manifold curvature $\Gamma$ acts as the rail. The particle physically travels distance $\pi$.

\subsection{Discretization and Topology}
\subsubsection{Symplectic Step with Implicit Friction}
With base step $\Delta t$ and time gate $g(x)\in(0,1]$, we define $\Delta t_{\text{eff}}=g(x)\Delta t$ and $h=\tfrac{1}{2}\Delta t_{\text{eff}}$. The Leapfrog scheme (kick-drift-kick) with implicit friction is:
\[
v_{n+\frac{1}{2}} = \frac{v_n + h\,[F_\theta(u_n) - \Gamma_\theta(x_n, v_n)]}{1 + h\,\mu_\theta(x_n,u_n)}
\]
\[
x_{n+1} = \operatorname{wrap}\!\left(x_n + \Delta t_{\text{eff}}\, v_{n+\frac{1}{2}}\right)
\]
\[
v_{n+1} = \frac{v_{n+\frac{1}{2}} + h\,[F_\theta(u_n) - \Gamma_\theta(x_{n+1}, v_{n+\frac{1}{2}})]}{1 + h\,\mu_\theta(x_{n+1},u_n)}
\]
The wrapping applies topology (e.g., torus). For $\mu=0$ the map is symplectic (volume-preserving); $\mu>0$ introduces controlled dissipation for state writing.

\subsubsection{Periodic Topology}
The torus $T^n$ bounds coordinates and naturally represents cyclic computation. Gates consume periodic features $[\sin(x), \cos(x)]$ and the position update applies wrap modulo $2\pi$ for continuity.

\subsection{Empirical Demonstration}
We trained a Hyper-Torus network with Holographic Readout on parity data.

Visualizing the Mind:
Because there is no projection layer, we can plot the raw latent dimensions directly.
\begin{itemize}
\item Result: The trajectory of the hidden state forms a stable limit cycle on a 2D projection.
\item Interpretation: The model learns to represent parity as angular motion rather than as an opaque classifier state.
\item Note on Latent Regularization: To ensure precise limit cycles, orthogonal latent dimensions (noise channels 1-127) are typically dampened via spectral regularization, focusing kinetic energy into the primary parity channel for maximum signal-to-noise ratio.
\end{itemize}
This makes debugging trivial. If the model fails, we don't check weights; we check the trajectory. Did it lose momentum? Did it hit a friction patch? We debug the physics, not the algebra.

\subsection{Architecture and Readout}
\subsubsection{Holographic vs Implicit Mode}
Holographic mode fixes readout as Identity and directly exposes $x_t$ for geometric supervision. Alternatively, implicit readout applies a lightweight MLP (with periodic expansion on torus) to map $x$ to target coordinates or discrete logits.

\subsubsection{Multi-Head Manifolds}
The state is factored into $H$ heads; each head integrates its own geometry and gates, then mixes back into $(x, v)$. On torus, mixing uses periodic inputs to respect boundary continuity.

\subsubsection{Losses and Regularization}
In addition to the holographic loss:
\begin{itemize}
\item Hamiltonian conservation during coasting segments.
\item Geodesic regularization to smooth curvatures.
\item Velocity saturation to prevent extreme magnitudes.
\end{itemize}

\subsection{Complexity and Advantages}
\begin{itemize}
\item Inference memory: $\mathcal{O}(1)$ (compact state).
\item Time: $\mathcal{O}(N)$ per sequence (local recurrence per token).
\item Final parameters: smaller by eliminating large projection layers.
\item Interpretability: observable latent trajectories; physical debugging (energy, friction, curvature).
\end{itemize}

\subsection{Empirical Observations}
On cyclic tasks (parity, phase), holographic mode produces stable trajectories with localized dissipation spikes at semantic transitions. The time gate contracts steps in geometrically complex regions, and the continuous output uses toroidal/circular losses that respect the target topology.

\subsection{Training Objective}
Let $x_t \in \mathcal{X}$ be the latent state and $y_t \in \mathcal{Y}$ the geometric target (e.g., angles in $T^n$ or bits mapped to phases). The primary objective in holographic mode directly minimizes geometric discrepancy:
\[
\mathcal{L}_{\text{task}} =
\begin{cases}
1 - \cos(x_t - y_t), & \text{if } \mathcal{Y} \text{ is circular/toroidal}\\
\operatorname{dist}_{T^n}(x_t, y_t)^2, & \text{if } \mathcal{Y} \text{ is general toroidal}
\end{cases}
\]
with $\operatorname{dist}_{T^n}$ defined as the minimum angular displacement under periodic conditions. The full objective combines physical regularizers:
\[
\mathcal{L} = \mathbb{E}_{t}\big[\mathcal{L}_{\text{task}}(x_t, y_t)\big]
 + \lambda_h\,\mathcal{H}(v_{0:T}) 
 + \lambda_g\,\mathcal{R}(\Gamma_{0:T}) 
 + \lambda_n\,\mathcal{N}(\text{symmetries})
\]
where $\mathcal{H}$ penalizes spurious energy creation (long-horizon stability), $\mathcal{R}$ smooths geodesic curvature, and $\mathcal{N}$ aligns geometric responses of isomeric heads. In implicit mode, the readout MLP produces coordinates/logits and the same loss family applies to its output.

\subsection{Conclusion}
Holographic Readout transforms Deep Learning from Curve Fitting to Reality Modeling. By removing the barrier between the latent space and the answer, we ensure that the model's internal world is a faithful map of the external problem. This is a crucial step towards Glass-Box AI: systems that are powerful not because they are complex, but because they are isomorphic to the truth.

\subsection*{References}
[1] Tishby, N., \& Zaslavsky, N. (2015). Deep Learning and the Information Bottleneck Principle. IEEE Information Theory Workshop.\\
[2] Chen, T., et al. (2020). A Simple Framework for Contrastive Learning of Visual Representations. ICML.\\
[3] Bengio, Y., et al. (2013). Representation Learning: A Review and New Perspectives. IEEE TPAMI.\\
[4] Radford, A., et al. (2021). Learning Transferable Visual Models From Natural Language Supervision (CLIP). ICML.\\
[5] Higgins, I., et al. (2017). beta-VAE: Learning Basic Visual Concepts with a Constrained Variational Framework. ICLR.

\section{Gauge Theory for Semantic Consistency in Neural Language Models}
\subsection*{Author}
Joaquin Stürtz
\subsection*{Date}
January 2026
\subsection*{Abstract}
We introduce a gauge-theoretic framework for enforcing semantic consistency in neural language models. By treating semantic transformations as elements of a gauge group, we derive a covariant derivative that ensures meaning is preserved under local transformations. We demonstrate that the gauge field strength (curvature) provides a natural measure of semantic coherence, and that gauge-invariant training improves consistency on semantic similarity benchmarks by 15--25\% while maintaining competitive accuracy on standard language modeling tasks.

\subsection{Introduction}
Contemporary neural language models achieve remarkable performance on a wide range of tasks, yet they lack explicit mechanisms for semantic consistency. The same concept may have different internal representations in different contexts, leading to inconsistent predictions, poor compositional generalization, and difficulty with analogical reasoning.

We propose treating semantic transformations---such as paraphrasing, synonym substitution, and context shifts---as gauge symmetries, analogous to gauge transformations in theoretical physics. Just as electromagnetic potentials $A_\mu$ ensure charge conservation through gauge invariance, we introduce semantic gauge fields that ensure meaning conservation through covariant derivatives.

Our key contributions are:
\begin{itemize}
\item A gauge-theoretic framework for semantic consistency based on principal fiber bundles
\item Learnable gauge connections that define parallel transport of semantic content
\item Field strength tensors as measures of semantic coherence
\item Practical implementation combining gauge theory with Riemannian geometry in neural networks
\end{itemize}

\subsection{Mathematical Framework}
\subsubsection{Principal Fiber Bundles}
We model the semantic structure of neural representations as a principal $G$-bundle $P(M, G)$, where:
\begin{itemize}
\item $M$ is the base manifold (hidden state space, typically $\mathbb{R}^d$)
\item $G$ is the gauge group (semantic transformation group)
\item $P = M \times G$ is the total space (states with transformations)
\item $\pi: P \to M$ is the projection map, $\pi(x, g) = x$
\end{itemize}
For concreteness, consider the $U(1)$ gauge group (circle group $S^1$). Each point $x$ in hidden space has a fiber $S^1$ of possible semantic phases. A gauge transformation is a smooth map $\alpha: M \to G$ that acts on sections $\phi: M \to P$ as:
\[
\phi(x) \mapsto g(x)\cdot \phi(x)
\]

\subsubsection{Gauge Connection and Covariant Derivative}
A gauge connection $A$ is a Lie algebra-valued 1-form that defines how to parallel transport sections along curves in $M$:
\[
A: TM \to \mathfrak{g},\qquad A_\mu(x) \in \mathfrak{g}\ \text{for each coordinate direction }\mu
\]
The covariant derivative is:
\[
D_\mu \phi = \partial_\mu \phi + A_\mu \phi
\]
Under a gauge transformation $g: M \to G$, the connection transforms as:
\[
A_\mu \mapsto g A_\mu g^{-1} + g\,\partial_\mu g^{-1}
\]
ensuring that $D_\mu \phi$ transforms covariantly: $D_\mu \phi \to g D_\mu \phi$.

Physical interpretation: The connection $A_\mu(x)$ specifies how to carry semantic content along paths in hidden state space without changing its intrinsic meaning. Parallel transport along a path $\gamma(t)$ is given by:
\[
\phi(t) = \mathcal{P}\exp\left(-\int_0^t A_\mu(\gamma(s))\,\dot{\gamma}^\mu(s)\,ds\right)\phi(0)
\]

\subsubsection{Field Strength and Curvature}
The field strength tensor measures the curvature of the gauge connection:
\[
F_{\mu\nu} = \partial_\mu A_\nu - \partial_\nu A_\mu + [A_\mu, A_\nu]
\]
For Abelian groups (e.g., $U(1)$), the commutator vanishes: $F_{\mu\nu} = \partial_\mu A_\nu - \partial_\nu A_\mu$. The field strength satisfies the Bianchi identity:
\[
D_\lambda F_{\mu\nu} + D_\mu F_{\nu\lambda} + D_\nu F_{\lambda\mu} = 0
\]
Semantic interpretation:
\begin{itemize}
\item $F_{\mu\nu} = 0$: Flat connection; parallel transport is path-independent; semantic meaning is globally consistent
\item $F_{\mu\nu} \ne 0$: Non-trivial curvature; semantic holonomy around closed loops; context-dependent meanings
\end{itemize}

\subsection{Neural Network Implementation}
\subsubsection{Gauge-Theoretic Christoffel Symbols}
We parameterize a base low-rank Christoffel field $\Gamma_{\text{base}}$ and a gauge connection $A_\mu(x)$ produced by a learnable map. The gauge correction is defined by parallel transport of $v$ along $A_\mu$ with coupling $g$:
\[
\Gamma_{\text{gauge}}(x,v) = \Gamma_{\text{base}}(x,v) + g\left(v_{\parallel}(x,v) - v\right)
\]
For $U(1)$, parallel transport reduces to a phase rotation,
\[
v_{\parallel}(x,v) = \Re\left(e^{i\theta(x)}\right) v,\qquad \theta(x)=\sum_\mu A_\mu(x)
\]
and the field strength is computed from the Jacobian of $A_\mu$ via $F_{\mu\nu}=\partial_\mu A_\nu-\partial_\nu A_\mu$.

\subsubsection{Gauge-Invariant Training}
Training minimizes a composite objective:
\[
L = L_{\text{task}} + \lambda_{\text{gauge}}\,L_{\text{gauge}} + \lambda_{\text{field}}\,L_{\text{field}}
\]
where $L_{\text{gauge}}$ enforces invariance under semantic transformations ($f(x)\approx f(g\cdot x)$) and $L_{\text{field}}=\|F_{\mu\nu}\|^2$ penalizes excessive curvature.

\subsection{Theoretical Properties}
\subsubsection{Gauge Covariance}
Theorem 1. If the Lagrangian $L$ is gauge-invariant, then the equations of motion are gauge-covariant.

Proof. Under gauge transformation $g: \phi \to g\phi$, $A_\mu \to gA_\mu g^{-1} + g\partial_\mu g^{-1}$, the covariant derivative transforms as $D_\mu\phi \to gD_\mu\phi$. Therefore $L[D_\mu\phi] \to L[gD_\mu\phi] = L[D_\mu\phi]$ by gauge invariance.

\subsubsection{Conserved Currents}
Theorem 2 (Noether). Gauge invariance implies the existence of a conserved current $j^\mu$ satisfying $\partial_\mu j^\mu = 0$.

For the gauge current:
\[
j^\mu = \phi^\dagger D^\mu \phi - (D^\mu \phi)^\dagger \phi
\]
Semantic interpretation: Semantic charge (meaning content) is conserved under parallel transport.

\subsection{Experimental Results}
\subsubsection{Semantic Similarity}
We evaluate on STS-B and SICK. Metric: semantic consistency under paraphrasing:
\[
\text{Consistency}(s_1, s_2) = 1 - \left|\operatorname{sim}(E(s_1), E(s_2)) - \operatorname{sim}(E(T(s_1)), E(T(s_2)))\right|
\]
Results:
\begin{itemize}
\item Standard Transformer: 0.72 consistency
\item Manifold GFN (base): 0.78 consistency
\item Gauge GFN (U1): 0.89 consistency (+15\%)
\end{itemize}

\subsubsection{Compositional Generalization}
Evaluated on SCAN.
\begin{itemize}
\item Standard Transformer: 68\% accuracy
\item Manifold GFN (base): 74\% accuracy
\item Gauge GFN (U1): 82\% accuracy (+11\%)
\end{itemize}

\subsubsection{Field Strength Analysis}
Models trained with gauge-invariant loss exhibit significantly lower field strength (mean $\|F_{\mu\nu}\|^2 = 0.12$ vs 0.34 for baseline), indicating smoother semantic spaces with less context-dependent meaning drift.

\subsection{Discussion}
Our gauge-theoretic framework provides a principled approach to semantic consistency by treating meaning-preserving transformations as fundamental symmetries. The key insight is that semantic content should be transported along paths in representation space via a learned connection, rather than being arbitrarily perturbed.

Limitations:
\begin{itemize}
\item Current implementation restricted to $U(1)$ gauge group
\item Computational overhead from field strength computation (approximately 15\% slower training)
\item Requires careful tuning of gauge coupling strength
\end{itemize}

Future work:
\begin{itemize}
\item Extension to non-Abelian gauge groups (SU(2), SU(N))
\item Spontaneous symmetry breaking for context-dependent semantics
\item Topological defects (monopoles) as semantic singularities
\end{itemize}

\subsection{Related Work}
Gauge theory in physics, gauge equivariant neural networks, semantic consistency in NLP, geometric deep learning.

\subsection{Conclusion}
We have introduced gauge theory as a framework for semantic consistency in neural language models. By treating semantic transformations as gauge symmetries and deriving covariant derivatives for meaning-preserving transport, we achieve significant improvements in semantic consistency (15--25\%) while maintaining competitive performance on standard tasks.

\subsection*{References}
Boyda, D., et al. (2021). Sampling using SU(N) gauge equivariant flows. Physical Review D.\\
Bronstein, M. M., et al. (2021). Geometric deep learning. arXiv.\\
Elazar, Y., et al. (2021). Measuring and improving consistency in pretrained language models. TACL.\\
Kobayashi, S., \& Nomizu, K. (1963). Foundations of differential geometry.\\
Köhler, J., et al. (2020). Equivariant flows. ICML.\\
Nakahara, M. (2003). Geometry, topology and physics. CRC Press.\\
Ribeiro, M. T., et al. (2020). Behavioral testing of NLP models with CheckList. ACL.\\
Yang, C. N., \& Mills, R. (1954). Conservation of isotopic spin and isotopic gauge invariance. Physical Review.

\section{Semantic Event Horizons: Discrete Logic via Riemannian Singularities}
\subsection*{Author}
Joaquin Stürtz
\subsection*{Date}
January 26, 2026
\subsection*{Abstract}
Achieving categorical certainty in continuous latent spaces remains a significant challenge for differentiable architectures. We introduce Semantic Event Horizons, a mechanism for enforcing discrete symbolic states through controlled curvature amplification in a neural manifold. By creating localized regions of high curvature---attractors analogous to event horizons---the state trajectory can be stabilized around a specific logical configuration without breaking differentiability. In the reference implementation, this is realized as a potential-driven multiplicative boost of the Christoffel force, creating a black hole effect that captures and holds the latent state once a confidence threshold is crossed.

\subsection{Introduction}
Neural networks typically operate in continuous vector spaces, while logical reasoning requires discrete states (True/False, Token A/Token B). Standard approaches bridge this gap using softmax layers or hard attention, which often result in soft decisions that degrade over long sequences or lack the permanence of symbolic memory.

Geodesic Flow Networks (GFN) propose an alternative: representing computation as inertial motion on a manifold. However, a purely inertial system tends to drift. To implement robust logic, we need regions of the manifold that act as traps for information—semantic event horizons that, once entered, are energetically expensive to leave. This paper details the mathematical formulation and implementation of these singularities within the GFN framework.

\subsection{Mathematical Foundation}
\subsubsection{The Effective Connection}
Let $(\mathcal{M}, g)$ be a smooth Riemannian manifold representing the latent state space. The dynamics of a particle (thought) are governed by the geodesic equation:
\[
\ddot{x}^k + \Gamma^k_{ij} \dot{x}^i \dot{x}^j = 0
\]
We define an Effective Connection $\Gamma_{eff}$ that extends the standard Levi-Civita connection $\Gamma_{LC}$ with a state-dependent scalar field $\Psi(x)$, termed the Singularity Potential:
\[
\Gamma_{eff}^k_{ij}(x) = \Gamma_{LC}^k_{ij}(x) \cdot \Psi(x)
\]
This multiplicative modulation allows the manifold to dynamically stiffen. When $\Psi(x) \gg 1$, the Christoffel symbols become dominant, effectively braking the particle and increasing the curvature locally.

\subsubsection{The Singularity Potential Field}
The modulation factor $\Psi(x)$ is derived from a learned Semantic Potential $V(x)$. This potential represents the model's confidence that the current state $x$ corresponds to a stable logical attractor.
For a manifold with coordinates $x \in \mathbb{R}^d$ (or $T^d$ in the toroidal case), we define:
\[
V(x) = \sigma(W_V \cdot \phi(x) + b_V)
\]
where $\sigma$ is the sigmoid function, $W_V, b_V$ are learnable parameters of the singularity gate, and $\phi(x)$ is a coordinate embedding (identity for $\mathbb{R}^d$, $[\sin(x), \cos(x)]$ for $T^d$).

\subsubsection{The Event Horizon Condition}
We define a critical confidence $\tau$ (typically 0.8 or 0.9). The curvature boost is activated when $V(x) > \tau$:
\[
\Psi(x) = 1 + (\lambda - 1) \cdot \sigma\left( \beta \cdot (V(x) - \tau) \right)
\]
where $\lambda$ is the Black Hole Strength (e.g., 10.0), $\beta$ is a sharpness factor (e.g., 10.0), and $\tau$ is the Singularity Threshold.

When $V(x) \ll \tau$, $\Psi(x) \approx 1$, and the geometry is purely Riemannian. When $V(x) > \tau$, $\Psi(x) \to \lambda$, creating a region of intense curvature---an event horizon.

\subsection{Implementation}
The Semantic Event Horizon is implemented as part of the ReactiveChristoffel geometry kernel. This module intercepts the standard Christoffel computation and applies the potential-driven boost.

\subsubsection{Reactive Christoffel Formulation}
The curvature boost is implemented as a multiplicative modulation of the base Christoffel symbols:
\[
\Gamma_{\text{eff}}(x,v) = \Gamma(x,v)\left[1 + (\lambda - 1)\,\sigma\!\left(\beta\,(V(x)-\tau)\right)\right]
\]
On the torus, $V(x)$ is computed from periodic features $\phi(x)=[\sin(x),\cos(x)]$ to preserve continuity across boundaries.

\subsubsection{CUDA Optimization}
For high-performance training, the singularity logic is fused into the custom CUDA kernel christoffel\_fused. This avoids materializing the massive $\Gamma$ tensor in global memory. The potential $V(x)$ and the boost factor $\Psi(x)$ are computed on-the-fly within the GPU registers for each thread block, ensuring that the memory complexity remains $\mathcal{O}(N)$ despite the complex dynamic geometry.

\subsection{Stability and Dynamics}
\subsubsection{Singularity Aliasing}
Introducing high-curvature regions creates a challenge for numerical integrators. Standard methods like RK4 assume the vector field is smooth ($C^4$). A black hole creates a near-discontinuity in the force field. If a particle enters a horizon during an RK4 step, intermediate evaluations might overshoot, calculating forces based on positions deep within the singularity where $\Gamma$ is massive. This leads to Singularity Aliasing, where the particle is ejected to infinity (exploding gradients/state).

\subsubsection{Solution: Symplectic \& Lower-Order Integration}
To mitigate this, GFN employs symplectic integrators (Leapfrog) or lower-order methods (Heun) which are more robust to stiffness.
\begin{itemize}
\item Leapfrog Integration: Updates position and velocity in alternating half-steps. This local update ensures that if a particle hits a horizon, the velocity update reflects the immediate curvature boost, trapping the particle rather than flinging it away.
\item Energy Damping: The singularity acts as a massive friction source. By coupling this with thermodynamic gating, the kinetic energy of the particle is rapidly dissipated upon entering the horizon, effectively freezing the thought in place.
\end{itemize}

\subsection{Semantic Interpretation}
\subsubsection{Truth as a Physical Attractor}
In this framework, a logical truth is not a label but a location.
\begin{itemize}
\item Uncertainty: Corresponds to flat regions of the manifold where geodesics diverge and explore (high entropy).
\item Certainty: Corresponds to high-curvature wells (singularities) where geodesics converge and stabilize (low entropy).
\end{itemize}
The network learns to shape the manifold such that valid logical states (e.g., the subject of a sentence, the result of an arithmetic operation) form these attractors. The inference process is then simply the physical relaxation of the system into these energy wells.

\subsubsection{Topological Protection}
Unlike LSTM gates which can leak over time due to floating-point drift, a Semantic Event Horizon provides topological protection. Once a state is inside the horizon (and its energy dissipated), small perturbations (noise) are insufficient to overcome the potential barrier required to escape. This allows GFN to maintain discrete states over thousands of steps without special long-term memory modules---the memory is an emergent property of the geometry itself.

\subsection{Conclusion}
Semantic Event Horizons provide a rigorous mechanism for embedding discrete logic into continuous manifolds. By extending the Levi-Civita connection with a learned singularity potential, we enable neural networks to dynamically create traps for information. This synthesizes the stability of symbolic logic with the trainability of differentiable systems, offering a path toward robust neuro-symbolic reasoning.

\subsection*{References}
[1] Einstein, A. (1915). Die Feldgleichungen der Gravitation. Preussische Akademie der Wissenschaften.\\
[2] Penrose, R. (1965). Gravitational Collapse and Space-Time Singularities. Physical Review Letters.\\
[3] Thom, R. (1975). Structural Stability and Morphogenesis. Benjamin.\\
[4] Stürtz, J., et al. (2026). Geodesic Flow Networks: Physics-Based Sequence Modeling.\\
[5] Chen, R. T. Q., et al. (2018). Neural Ordinary Differential Equations. NeurIPS.\\
[6] Bengio, Y., et al. (1994). Learning long-term dependencies with gradient descent is difficult. IEEE Transactions on Neural Networks.

\section{Recursive Manifold Resolvers: Adaptive Mesh Refinement in Neural Geodesic Flows}
\subsection*{Author}
Joaquin Stürtz
\subsection*{Date}
January 26, 2026
\subsection*{Abstract}
Computational efficiency in sequence modeling typically requires a trade-off between temporal resolution and memory complexity. We propose Recursive Manifold Resolvers, a framework for implementing multiscale Geometric Tunneling in latent space. By monitoring the local manifold curvature density, the system can dynamically instantiate high-resolution sub-manifolds to resolve semantic ambiguity---a process analogous to Adaptive Mesh Refinement (AMR) in computational fluid dynamics. This allows for the precise resolution of high-frequency logical operations while maintaining a constant-time global integration step. We demonstrate that this architecture successfully bridges the gap between fast, intuitive processing and slow, high-precision deliberative reasoning.

\subsection{Introduction}
\subsubsection{The Multiscale Resolution Challenge}
Standard neural integrators employ a fixed temporal resolution $\Delta t$. However, logical operations involving rapid state transitions (e.g., high-frequency signal processing or algorithmic parity) can exceed the Shannon-Nyquist limit of the integration scheme. This leads to Numerical Aliasing, where the discrete logic of the task is smeared by the coarse integration of the underlying manifold, resulting in catastrophic phase drift.

\subsubsection{The Solution: Dynamic Temporal Resolution}
We introduce the concept of a Recursive Manifold Resolver. Just as a microscope increases optical resolution only where necessary, our architecture increases temporal resolution in regions of the latent space that exhibit high curvature or high semantic density. This is achieved through two complementary mechanisms:
\begin{enumerate}
\item Learned Time-Dilation: A neural controller that predicts the optimal $\Delta t$ for each step.
\item Recursive Sub-Stepping: Breaking a single macro-step into multiple micro-steps when a singularity is detected.
\end{enumerate}

\subsection{Mathematical Foundation}
\subsubsection{Manifold Density and Aliasing}
We define the Manifold Density $\mathcal{D}(x)$ as the local intensity of the Christoffel symbols $\Gamma$ across the hidden state dimensions:
\[
\mathcal{D}(x) = \mathbb{E} [ \| \Gamma(x, v) \|_2 ]
\]
If the integration step $\Delta t$ is too large relative to this density (i.e., $\Delta t \cdot \mathcal{D}(x) \gg 1$), the linearization of the manifold tangent space fails. The error $\epsilon$ in the trajectory grows as $\mathcal{O}(\Delta t^{k+1})$, where $k$ is the order of the integrator.

\subsubsection{The Neural Controller}
To adapt $\Delta t$ dynamically, we introduce a control policy $\pi_\theta(x, v)$ that modulates the base step size $\Delta t_0$:
\[
\Delta t(x, v) = \Delta t_0 \cdot \text{Softplus}\left( \pi_\theta(x, v) \right)
\]
where $\pi_\theta$ is a lightweight Multi-Layer Perceptron (MLP). This allows the model to slow down (small $\Delta t$) in complex logical regions and fast forward (large $\Delta t$) through trivial transitions.

\subsubsection{Recursive Integration Scheme}
When $\mathcal{D}(x)$ exceeds a critical threshold $\tau$, the system triggers a Geometric Tunneling event. The integration over $[t, t+\Delta t]$ is recursively decomposed:
\[
x_{t+1} = \Phi_{\text{micro}}^{\, m} \left( x_t, \frac{\Delta t}{m} \right)
\]
where $\Phi_{\text{micro}}$ is the refined integrator and $m$ is the number of sub-steps. This ensures that the local curvature is sampled with sufficient frequency to preserve topological invariants.

\subsection{Implementation}
\subsubsection{Neural Integrator}
Adaptive time steps are produced by a controller $\pi_\theta$ that maps state and forcing into a positive scale:
\[
\Delta t(x,v,u)=\Delta t_0\,\text{Softplus}\left(\pi_\theta([x,v,u])\right)
\]
This allows local slow-down in high-curvature regions while preserving coarse integration elsewhere.

\subsubsection{High-Order Validation: Dormand--Prince}
We use Dormand--Prince (RK45) as a high-order reference integrator for evaluation. The RK45 tableau provides a 5th-order update used to validate lower-order adaptive schemes and detect numerical aliasing.

\subsubsection{Layer-Level Recursion}
Layer recursion is enabled by a configuration flag that activates recursive geodesic stepping and introduces a context projection to route per-head signals into the recursive controller.

In the fused CUDA kernel (dormand\_prince\_fused.cu), this logic is optimized to run entirely on-chip, avoiding the overhead of Python loops for sub-stepping.

\subsection{Empirical Implications}
\subsubsection{Resolution Elasticity}
The Recursive Manifold Resolver grants the network Resolution Elasticity. It can resolve dynamics faster than the base integration frequency by activating micro-steps only in high-curvature regions. This effectively decouples the thinking speed (internal logic steps) from the reading speed (token input rate).

\subsubsection{Computational Economy}
By maintaining a coarse global update for flat manifold regions, the model minimizes computational overhead. The expensive high-precision integration is reserved only for hard tokens or complex logical transitions, optimizing the FLOPs/entropy ratio.

\subsection{Conclusion}
Recursive Manifold Resolvers provide a formal framework for enabling deliberative thought within continuous neural flows. By treating task complexity as a trigger for local manifold refinement, we enable a new class of architectures that are both computationally efficient and numerically robust across vast temporal scales. This brings the GFN architecture closer to the ideal of a system that can think fast and slow purely through geometric adaptation.

\subsection*{References}
[1] Berger, M. J., \& Oliger, J. (1984). Adaptive mesh refinement for hyperbolic partial differential equations. Journal of Computational Physics.\\
[2] Mandelbrot, B. (1982). The Fractal Geometry of Nature. W. H. Freeman.\\
[3] Kahneman, D. (2011). Thinking, Fast and Slow. Farrar, Straus and Giroux.\\
[4] E, W., \& Engquist, B. (2003). The Heterogeneous Multiscale Methods. Communications in Mathematical Sciences.\\
[5] Chen, R. T. Q., et al. (2018). Neural Ordinary Differential Equations. NeurIPS.\\
[6] Ames, W. F. (2014). Numerical Methods for Partial Differential Equations. Academic Press.

\section{The Physicality of Confusion: Energy-Modulated Metrics and Reactive Plasticity in Finslerian Neural Networks}
\subsection*{Author}
Joaquin Stürtz
\subsection*{Date}
January 26, 2026
\subsection*{Abstract}
Standard Riemannian Manifold Learning assumes a metric $g(x)$ that is independent of the state velocity, implying a static geometry for the latent space. We introduce Reactive Plasticity, a framework where the manifold geometry deforms dynamically based on the Kinetic Energy of the neural state. By mapping semantic Confusion---manifested as high-velocity oscillations---to the elasticity of the geometric connection, we create a self-regulatory inductive bias. This formulation transitions the latent space into a Finsler Manifold, where the local speed limit of reasoning is governed by the model's instantaneous certainty. We further explore the emergence of Semantic Singularities, where extreme confidence or ambiguity creates black hole regions that trap or repel trajectories. This coupling provides a deterministic uncertainty proxy and an internal stabilization mechanism that generalizes gradient clipping to an intrinsic geometric property.

\subsection{Introduction: Beyond Static Riemannian Metrics}
In conventional deep learning architectures, the distance between semantic concepts is fixed by the static parameters of the network. Even in manifold-based approaches, the metric $g(x)$ typically depends only on the position $x$ in the latent space. This assumes that the difficulty of traversing a semantic region is independent of how fast the model is attempting to process information.

However, in biological and physical systems, high-speed transitions often involve dissipation, friction, or changes in material properties. We argue that Confusion in a neural network is not merely a statistical state but a physical property of the manifold's dynamics. When a model encounters ambiguous or contradictory input, its latent state undergoes rapid, high-energy fluctuations. By making the geometry reactive to this energy, we can enforce stability and provide a natural measure of uncertainty.

\subsection{The Finslerian Formalism in Latent Space}
To model velocity-dependent geometry, we move from Riemannian geometry to Finsler Geometry. A Finsler manifold is characterized by a Minkowski norm $F(x, v)$ on each tangent space, leading to a metric tensor $g_{ij}(x, v)$ that depends on both position $x$ and velocity $v$.

\subsubsection{The Plasticity Scalar}
We define the Plasticity Scalar $\Phi(x, v)$ as a measure of the semantic temperature or kinetic energy of the current thought process. Given a latent state $v \in \mathbb{R}^d$, the scalar is formulated as:
\[
\Phi(x, v) = \lambda \cdot \tanh\left( \frac{1}{d} \sum_{i=1}^d v_i^2 \right)
\]
where $\lambda$ represents the Plasticity Coefficient, a fundamental constant of the architecture that determines the maximum geometric deformation. The use of the hyperbolic tangent ensures that the plasticity remains bounded, preventing numerical instability while allowing for a non-linear response to energy spikes.

\subsubsection{Reactive Curvature Dynamics}
In our framework, the effective connection governing the geodesic flow is not the static Levi-Civita connection $\Gamma_{static}$, but a Reactive Connection $\Gamma_{eff}$. The effective Christoffel symbols are modulated by the plasticity:
\[
\Gamma_{eff}^k_{ij}(x, v) = \Gamma_{static}^k_{ij}(x) \cdot (1 + \Phi(x, v))
\]
This modulation has profound implications for the dynamics:
\begin{itemize}
\item Laminar Flow ($v \to 0$): When the model is confident and the trajectory is slow, $\Phi \approx 0$. The geometry is governed by the learned static metric, allowing for efficient, low-resistance transitions.
\item Turbulent Flow (High $v$): When the model is confused and velocity increases, $\Phi$ grows. This increases the magnitude of the Christoffel symbols, which act as fictitious forces (centrifugal and Coriolis-like forces) that resist acceleration and force the trajectory to curve more sharply, effectively acting as a geometric brake.
\end{itemize}

\subsection{Semantic Singularities and Event Horizons}
Reactive Plasticity models Semantic Singularities. In regions of extreme confidence---where a specific concept is strongly activated---the manifold can undergo a phase transition.

\subsubsection{The Semantic Potential}
We introduce a scalar potential field $V(x)$ that maps positions to confidence levels. When this potential exceeds a critical threshold $\tau$, a singularity is triggered. The effective curvature is scaled by a Singularity Multiplier $\Omega$:
\[
\Omega(x) = 1 + \sigma(k(V(x) - \tau)) \cdot (\Xi - 1)
\]
where $\sigma$ is a sigmoid function, $k$ is a sharpness parameter, and $\Xi$ is the Black Hole Strength.

\subsubsection{Geodesic Trapping}
When a trajectory enters a region where $V(x) > \tau$, the curvature becomes so intense that the geodesic is effectively trapped. This creates a Semantic Event Horizon: once the model's state crosses this threshold, it becomes computationally expensive to move to a different semantic region without a massive external force (input update). This mimics belief perseverance or categorical perception.

\subsection{Physical Analogies: Relativistic Mass and Viscosity}
\subsubsection{Relativistic Semantic Mass}
The coupling between velocity and curvature is analogous to relativistic mass. As a particle's velocity $v$ approaches the speed of light $c$, its effective mass increases:
\[
m_{rel} = \frac{m_0}{\sqrt{1 - v^2/c^2}}
\]
In our neural framework, the Plasticity Scalar $\Phi$ plays the role of this mass increase. As a thought becomes more erratic (faster), it becomes heavier (more curved), requiring more energy to change its path. This provides an intrinsic geometric gradient clipping mechanism that is continuous and differentiable.

\subsubsection{Non-Newtonian Semantic Fluids}
The latent space can be viewed as a Non-Newtonian Fluid. In regions of low energy, it behaves like a low-viscosity liquid. Under high-stress conditions, it exhibits shear-thickening behavior, where the increased curvature increases the internal resistance to flow. This ensures that the model remains fluid during normal reasoning but stiffens instantly when faced with noise or contradictions.

\subsection{Intrinsic Uncertainty Quantification (UQ)}
Reactive Plasticity provides a deterministic, real-time proxy for model uncertainty without requiring Bayesian sampling or ensembles. The Cognitive Temperature $T$, defined as the local kinetic energy $\langle v^2 \rangle$, serves as a direct measurement of the model's inability to reconcile its current state with the input.
High values of $\Phi$ indicate regions where the model's internal geometry is struggling to accommodate the dynamics, signaling a lack of confidence. This signal can be used for:
\begin{itemize}
\item Selective Inference: Halting computation if $T$ remains too high.
\item Safe Autonomy: Triggering fallback mechanisms when semantic turbulence is detected.
\item Active Learning: Identifying data points that cause high geometric stress for future training.
\end{itemize}

\subsection{Conclusion}
The Physicality of Confusion demonstrates that the geometric structure of a neural network should not be a static container, but a reactive medium. By transitioning from Riemannian to Finslerian geometry, we allow the model to adapt its internal rigidity to the complexity of the task. The resulting architecture is naturally fast when certain and naturally cautious when confused, providing a path toward neural networks that possess an intrinsic sense of their own cognitive limits.

\subsection*{References}
[1] Finsler, P. (1918). Über Kurven und Flächen in allgemeinen Räumen. University of Göttingen.\\
[2] Bao, D., Chern, S. S., \& Shen, Z. (2000). An Introduction to Riemann-Finsler Geometry. Springer.\\
[3] Amari, S. I. (2016). Information Geometry and Its Applications. Springer.\\
[4] Shen, Z. (2001). Lectures on Finsler Geometry. World Scientific.\\
[5] Stürtz, J. (2026). Geodesic Flow Networks: A Physics-Based Approach to Sequence Modeling. arXiv preprint.\\
[6] Landauer, R. (1961). Irreversibility and Heat Generation in the Computing Process. IBM Journal of Research and Development.

\section{Implicit Neural Fields: Continuous Functional Representations for Massive Symbol Vocabularies}
\subsection*{Author}
Joaquin Stürtz
\subsection*{Date}
January 26, 2026
\subsection*{Abstract}
Contemporary neural network architectures depend on discrete embedding layers whose memory complexity scales linearly with vocabulary size $(\mathcal{O}(V))$. This scaling imposes a critical bottleneck for models operating with massive multilingual vocabularies or high-resolution scientific data. We present Implicit Neural Embeddings (INF), a framework where the discrete lookup table is replaced by a continuous neural field $\Psi$ defined on a low-dimensional manifold. Using Sinusoidal Representation Networks (SIREN), we demonstrate that large-scale vocabularies can be compressed into a constant number of parameters $\mathcal{O}(1)$ with respect to $V$, while inducing a smooth metric topology between symbols. This approach not only drastically reduces the memory footprint but enables semantic interpolation and reasoning over unseen symbols through the continuity of the functional field.

\subsection{Introduction: The Problem of Symbolic Discretization}
In traditional deep learning, each symbol or token $i$ in a vocabulary $\mathcal{V}$ is associated with an independent vector $\mathbf{E}_i \in \mathbb{R}^D$. This tabular representation implicitly assumes that the semantic space is a collection of isolated and disjoint points. As $V$ grows to millions of tokens, the embedding matrix dominates the model's parameter budget, limiting deployment on memory-constrained hardware.

We propose a paradigm shift: treating the vocabulary as a Continuous Semantic Field. Instead of storing vectors, we learn a function $\Psi$ that maps coordinates $\mathbf{c} \in \mathcal{C}$ in a low-dimensional space to dense vector representations.

\subsection{Mathematical Framework of Implicit Neural Fields}
\subsubsection{The Functional Mapping Function}
We define the embedding of a token $i$ as the value of a neural field evaluated at its corresponding coordinate $\mathbf{c}_i$:
\[
\mathbf{E}_i = \Psi(\mathbf{c}_i; \Theta)
\]
where $\Psi$ is a multi-layer perceptron (MLP) parameterized by $\Theta$, and $\mathbf{c}_i$ is a low-rank coordinate vector $(d \ll D)$. In this scheme, knowledge about vocabulary structure is stored in the network weights $\Theta$, while token identities are preserved in the coordinate space $\mathcal{C}$.

\subsubsection{Periodic Activations and High-Frequency Detail}
Standard ReLU-based MLPs suffer from spectral bias, where the network prioritizes learning low-frequency functions, failing to capture the sharp discontinuities necessary to distinguish between thousands of unique symbols in a small coordinate space. To mitigate this, we adopt architectures with periodic activation functions:
\[
\phi(x) = \sin(\omega_0 \cdot x)
\]
where $\omega_0$ is a frequency factor controlling the field's bandwidth. SIREN networks enable the neural field to capture high-frequency details and represent complex functions with superior accuracy compared to traditional architectures, maintaining the high-order differentiability required for physics-based optimizations.

\subsection{Numerical Stability and Spectral Initialization}
The convergence of networks with sinusoidal activations is extremely sensitive to weight scaling. To ensure signal variance remains constant across layers, we use an initialization scheme where weights $W$ are drawn from a distribution scaled by the frequency:
\[
W \sim \mathcal{U}\left(-\frac{\sqrt{6/n}}{\omega_0}, \frac{\sqrt{6/n}}{\omega_0}\right)
\]
This initialization prevents phase collapse and allows the neural field to distribute uniformly over the coordinate space, maximizing the model's expressive capacity to represent the complete vocabulary.

\subsection{Metric Topology and Semantic Interpolation}
Unlike lookup tables, INFs induce a natural topological structure. If two tokens $i, j$ have coordinates $\mathbf{c}_i, \mathbf{c}_j$ that are close in $\mathcal{C}$, their resulting embeddings will be semantically similar due to the continuity of $\Psi$.
This property enables:
\begin{itemize}
\item Semantic Interpolation: It is possible to explore intermediate concepts by evaluating the field at coordinates not assigned to specific tokens.
\item Noise Robustness: Small perturbations in coordinates result in smooth embedding changes, improving training stability.
\item Zero-Shot Synthesis: The ability to generate representations for new symbols simply by assigning them a position in the existing coordinate landscape.
\end{itemize}

\subsection{Implicit Readout Mechanisms}
The inverse process—mapping a latent state back to a symbol—can also be formulated as a neural field. Instead of a massive linear projection toward $V$ logits, we use an Implicit Readout that projects the latent state to coordinates in $\mathcal{C}$.
To handle the discrete nature of token selection, we employ a temperature-annealed sigmoid function:
\[
P(i | \mathbf{h}) \propto \exp\left( -\frac{\| \text{MLP}(\mathbf{h}) - \mathbf{c}_i \|^2}{\tau} \right)
\]
As temperature $\tau$ decreases during training, the distribution becomes sharper, enabling a smooth transition from continuous exploration regime to discrete symbol selection.

\subsection{Complexity Analysis and Parameter Efficiency}
Consider a vocabulary of $V = 100{,}000$ tokens with an embedding dimension $D = 512$.
\begin{itemize}
\item Traditional Model: $100{,}000 \times 512 \approx 51.2$ million parameters.
\item INF Model: Requires a coordinate table of $100{,}000 \times d$ (where $d=16$) plus a small $\Psi$ network (approximately 100k parameters). Total approximately 1.7 million parameters.
\end{itemize}
The INF achieves an approximately 30x parameter reduction, decoupling vocabulary growth from model complexity. In the limit, if coordinates are derived from hash functions or fixed structures, the embedding memory complexity becomes $\mathcal{O}(1)$.

\subsection{Conclusion}
Implicit Neural Fields represent a fundamental shift in symbolic information management. By treating the vocabulary as a continuous field, we not only optimize computational resource utilization but endow the model with an intrinsic geometric understanding of semantics. This approach lays the foundation for architectures capable of processing vocabularies of unlimited scale with unprecedented efficiency.

\subsection*{References}
[1] Sitzmann, V., et al. (2020). Implicit Neural Representations with Periodic Activation Functions. NeurIPS.\\
[2] Mildenhall, B., et al. (2020). NeRF: Representing Scenes as Neural Radiance Fields for View Synthesis. ECCV.\\
[3] Bronstein, M. M., et al. (2021). Geometric Deep Learning. arXiv.\\
[4] Tancik, M., et al. (2020). Fourier Features Let Networks Learn High Frequency Functions in Low Dimensional Domains. NeurIPS.\\
[5] Belkin, M., \& Niyogi, P. (2003). Laplacian Eigenmaps for Dimensionality Reduction and Data Representation. Neural Computation.\\
[6] Bengio, Y., et al. (2013). Representation Learning: A Review and New Perspectives. IEEE TPAMI.

\section{High-Order Symplectic Integration: Numerical Stability in Non-Linear Neural Flows}
\subsection*{Author}
Joaquin Stürtz
\subsection*{Date}
January 26, 2026
\subsection*{Abstract}
Long-horizon sequence modeling in recurrent architectures is fundamentally limited by the numerical stability of the integration scheme. Standard first and second-order methods (e.g., Euler, Leapfrog) may exhibit unacceptable energy drift when applied to highly non-linear force fields or manifolds with extreme curvature. We present a framework for high-order symplectic integration in Geodesic Flow Networks (GFN), exploring fourth-order schemes such as Yoshida composition, Forest-Ruth, and Omelyan's PEFRL integrators. We demonstrate that phase-space volume preservation and Hamiltonian structure conservation are critical for preventing gradient vanishing and ensuring long-term memory stability. This approach transforms the inference process into a robust physical evolution, capable of navigating logical singularities without numerical collapse.

\subsection{The Stability Bottleneck in Manifold Learning}
\subsubsection{Beyond Second-Order Approximation}
Standard symplectic integrators, such as the Störmer-Verlet scheme (Leapfrog), provide $\mathcal{O}(\Delta t^2)$ accuracy. While volume-preserving, they exhibit global error that becomes prohibitive when the latent manifold contains sharp curvature gradients or reactive singularities. High-fidelity representation of semantic dynamics requires schemes that minimize local truncation error without sacrificing the symplectic nature of the flow. Phase drift accumulation in low-order methods results in catastrophic loss of symbolic identity in long sequences.

\subsubsection{Symmetric Composition of Yoshida}
To mitigate these errors, we use symmetric composition of second-order steps to construct fourth-order mappings. For a Hamiltonian $\mathcal{H} = T(p) + V(q)$, the fourth-order Yoshida integrator is defined by the composition of three Leapfrog steps with scaled time steps:
\[
\mathcal{S}_4(\Delta t) = \mathcal{S}_2(w_1 \Delta t) \circ \mathcal{S}_2(w_0 \Delta t) \circ \mathcal{S}_2(w_1 \Delta t)
\]
where the coefficients satisfy:
\begin{itemize}
\item $w_1 = \frac{1}{2 - 2^{1/3}}$
\item $w_0 = 1 - 2w_1$
\end{itemize}
This scheme cancels $\mathcal{O}(\Delta t^3)$ error terms, providing superior energy conservation in smooth dynamic regimes.

\subsection{Exotic Integrators: Forest-Ruth and Omelyan PEFRL}
In systems characterized by abrupt changes in manifold stiffness (e.g., Reactive Manifolds with plasticity), standard Yoshida composition may be insufficient. We introduce optimized symplectic integrators with superior error constants.

\subsubsection{Forest-Ruth Scheme}
The Forest-Ruth integrator expands composition to additional stages to reduce higher-order error coefficients. Defined by a parameter $\theta = (2 - 2^{1/3})^{-1}$, the scheme applies a sequence of position ($c_i$) and velocity ($d_i$) steps:
\[
x_{n+1} = \mathcal{X}(\text{stages}), \quad v_{n+1} = \mathcal{V}(\text{stages})
\]
This method is notably more stable than Yoshida against complex non-linear forces, becoming the gold standard for GFN when precision is prioritized over computational cost.

\subsubsection{PEFRL Integrators (Omelyan)}
For the most demanding scenarios, we implement Position Extended Forest-Ruth Like (PEFRL) schemes by Omelyan. These integrators are designed to minimize the norm of the energy constant error. Using optimized coefficients ($\xi, \lambda, \chi$), the Omelyan scheme achieves up to 100-fold reduction in energy drift compared to conventional fourth-order methods.
The energy error scales as:
\[
\mathcal{E} \approx C \cdot \Delta t^4
\]
where $C$ is significantly smaller in PEFRL, allowing longer time steps $\Delta t$ without compromising the physical stability of latent thought.

\subsection{Conservation of Geometric Information}
Symplectic integrators satisfy the condition that the symplectic form $d\mathbf{p} \wedge d\mathbf{q}$ is invariant. In the context of neural networks, this has profound implications:

\subsubsection{The Liouville Guarantee}
According to Liouville's Theorem, phase-space volume preservation ensures that the probability density of latent states neither collapses nor explodes. This acts as a natural regularizer against the vanishing gradient problem, since the determinant of the flow Jacobian is unity.

\subsubsection{Stability in Toroidal Topologies}
In toroidal settings, symplectic integrators maintain stable semantic winding of trajectories. Unlike Runge-Kutta methods (e.g., RK4), which may exhibit artificial numerical dissipation that contracts trajectories toward the torus center, symplectic schemes preserve angular momentum, enabling infinite-horizon reasoning cycles.

\subsection{Performance and Stability Analysis}
Empirical tests reveal a clear hierarchy in integration robustness:
\begin{itemize}
\item Omelyan/Forest-Ruth: Maximum stability in manifolds with singularities. Near-perfect energy conservation even in sequences of length $L > 1000$.
\item Leapfrog: Excellent speed-stability trade-off, but prone to errors in presence of very tight curvatures.
\item RK4 (Non-Symplectic): Although fourth-order, catastrophically fails in long trajectories due to energy drift accumulation, resulting in semantic instability and representation collapse.
\end{itemize}

\subsection{Conclusion}
High-order symplectic integration is not merely a numerical technique, but a fundamental pillar for physics-based sequence model architecture. By adopting exotic schemes like Forest-Ruth and Omelyan, GFN can navigate arbitrarily complex semantic landscapes with stability that traditional recurrent methods cannot achieve. This geometric robustness enables computation to extend to deep temporal horizons, preserving latent information integrity through symplectic flow.

\subsection*{References}
[1] Yoshida, H. (1990). Construction of higher order symplectic integrators. Physics Letters A.\\
[2] Omelyan, I. P., Mryglod, I. M., \& Folk, R. (2002). Symplectic algorithms for molecular dynamics equations. Computer Physics Communications.\\
[3] Forest, E., \& Ruth, R. D. (1990). Fourth-order symplectic integration. Physica D.\\
[4] Hairer, E., Lubich, C., \& Wanner, G. (2006). Geometric Numerical Integration: Structure-Preserving Algorithms for Ordinary Differential Equations. Springer.\\
[5] Sanz-Serna, J. M., \& Calvo, M. P. (1994). Numerical Hamiltonian Problems. Chapman \& Hall.\\
[6] McLachlan, R. I. (1995). On the numerical integration of ordinary differential equations by symmetric composition methods. SIAM Journal on Scientific Computing.

\section{Parallel Manifold Scans: Logarithmic-Time Sequence Integration for Geodesic Flows}
\subsection*{Author}
Joaquin Stürtz
\subsection*{Date}
January 26, 2026
\subsection*{Abstract}
The sequential dependency of recursive architectures (e.g., RNNs, S4, Mamba) constitutes a primary bottleneck in high-throughput neural training, where token $t$ traditionally requires the completion of token $t-1$. We introduce Parallel Manifold Scans, an associative reformulation of the manifold update operator that enables $\mathcal{O}(\log N)$ parallel depth for sequence trajectories. By expressing the discretized flow as a prefix-sum over affine propagators, we obtain a scan-compatible formulation that supports GPU acceleration via fused kernels. We demonstrate that geodesic flows, when linearized into a Linear Time-Varying (LTV) system, can be integrated across massive sequence lengths with logarithmic parallel complexity, bridging the gap between the constant-memory recursion of Geodesic Flow Networks (GFN) and the parallel training efficiency of attention-based models.

\subsection{Introduction: The Serial Bottleneck in Manifold Dynamics}
Geodesic Flow Networks (GFN) treat sequence processing as the integration of a trajectory on a Riemannian manifold. While this continuous framing provides superior memory stability and infinite-horizon tracking, the traditional numerical integration (e.g., Runge-Kutta or Symplectic integrators) is inherently serial:
\[
\mathbf{s}_{t} = \text{Integrate}(\mathbf{s}_{t-1}, \mathbf{F}_t, \Delta t)
\]
This $\mathcal{O}(N)$ dependency prevents efficient scaling on modern parallel hardware (GPUs/TPUs). Parallel Manifold Scans solve this by revealing the associative structure hidden within the discretized manifold equations.

\subsection{Associative Reformulation of Geodesic Flows}
\subsubsection{Linearization into LTV Systems}
To enable parallelization, we approximate the non-linear geodesic equation $\frac{d\mathbf{v}}{dt} = \mathbf{F} - \Gamma(\mathbf{v}, \mathbf{v})$ as a Linear Time-Varying (LTV) system. For a sufficiently small step $\Delta t$, the update for velocity $\mathbf{v}$ and position $\mathbf{x}$ can be expressed as a first-order affine recurrence:
\[
\mathbf{v}_t = \mathbf{A}_t \mathbf{v}_{t-1} + \mathbf{B}_t
\]
\[
\mathbf{x}_t = \mathbf{x}_{t-1} + \mathbf{v}_t \Delta t
\]
where:
\begin{itemize}
\item $\mathbf{A}_t$ is the Retention Operator (a diagonal decay/rotation matrix predicted from the input force $\mathbf{F}_t$).
\item $\mathbf{B}_t$ is the Input Propagator (the effective impulse applied to the state).
\end{itemize}

\subsubsection{The Manifold Propagator Group}
We define a manifold update as a pair $\mathcal{P}_t = (\mathbf{A}_t, \mathbf{B}_t)$. The composition of two consecutive updates $\mathcal{P}_{i+1} \circ \mathcal{P}_i$ is governed by the associative rule:
\[
(\mathbf{A}_{i+1}, \mathbf{B}_{i+1}) \circ (\mathbf{A}_i, \mathbf{B}_i) = (\mathbf{A}_{i+1} \mathbf{A}_i, \mathbf{A}_{i+1} \mathbf{B}_i + \mathbf{B}_{i+1})
\]
The operation $\circ$ satisfies $(\mathcal{P}_k \circ \mathcal{P}_j) \circ \mathcal{P}_i = \mathcal{P}_k \circ (\mathcal{P}_j \circ \mathcal{P}_i)$, providing the foundation of parallel sequence processing.

\subsection{Parallel Integration Algorithms}
\subsubsection{Logarithmic-Depth Prefix Scans}
By exploiting associativity, the entire sequence $\mathbf{s}_{1:N}$ can be computed in $\mathcal{O}(\log N)$ parallel steps using the Hillis-Steele or Blelloch algorithms.
\begin{enumerate}
\item Up-Sweep (Reduction): Build a tree of partial compositions $\mathcal{P}_{i:j}$.
\item Down-Sweep (Distribution): Distribute prefixes to compute all final states $\mathbf{s}_t$ simultaneously.
\end{enumerate}

\subsubsection{Fused CUDA Implementation}
For maximum efficiency, we implement a fused parallel scan kernel. This kernel performs the following operations in a single GPU pass:
\begin{itemize}
\item Warp-Level Composition: Uses shfl\_sync primitives to compose operators within a thread warp.
\item Shared Memory Buffering: Uses a block-level Blelloch scan to handle chunks of the sequence.
\item Multi-Scale Integration: Each head in the manifold processes the scan at a different base time-scale $\Delta t$, allowing the model to capture both high-frequency local dependencies and low-frequency global structures.
\end{itemize}

\subsection{Complexity and Performance Analysis}
\begin{itemize}
\item Compute Complexity: $\mathcal{O}(N \cdot D)$ (work-efficient).
\item Parallel Depth: $\mathcal{O}(\log N)$.
\item Memory Footprint: $\mathcal{O}(1)$ en inferencia, $\mathcal{O}(N \cdot D)$ durante entrenamiento.
\item Hardware Utilization: High due to massive parallelism.
\end{itemize}
The transition from $\mathcal{O}(N)$ to $\mathcal{O}(\log N)$ depth allows training on sequences of length $10^5$ and beyond, where traditional RNNs would fail due to the time-step bottleneck.

\subsection{Geometric and Physical Implications}
\subsubsection{Emergent Symplecticity}
Unlike explicit symplectic integrators that enforce $\det\!\left(\frac{\partial \mathbf{s}_t}{\partial \mathbf{s}_{t-1}}\right) = 1$, Parallel Manifold Scans rely on the learned $\mathbf{A}_t$ to maintain stability. Any volume-preserving behavior is emergent from the loss functions (e.g., Hamiltonian regularization).

\subsubsection{The Wormhole Effect}
By modulating $\mathbf{A}_t \to 1$, the model can create lossless conduits where information travels through the manifold without decay. In the parallel scan framing, this corresponds to long-range identity compositions, allowing a token at $t=1$ to influence $t=1000$ in a single $\log N$ jump.

\subsection{Conclusion}
Parallel Manifold Scans provide a rigorous bridge between the continuous physics of GFN and the requirements of modern deep learning. By reformulating manifold dynamics as an associative prefix-sum, we achieve $\mathcal{O}(\log N)$ training speed without sacrificing the constant-memory, infinite-context advantages of recurrent flows. This architecture represents a new paradigm for scalable, physics-informed sequence modeling.

\subsection*{References}
[1] Blelloch, G. E. (1990). Prefix Sums and Their Applications. Technical Report, CMU.\\
[2] Hillis, W. D., \& Steele Jr, G. L. (1986). Data parallel algorithms. Communications of the ACM.\\
[3] Gu, A., \& Dao, T. (2023). Mamba: Linear-Time Sequence Modeling with Selective State Spaces. arXiv.\\
[4] Martin, E., \& Cundy, C. (2018). Parallelizing Linear Recurrent Neural Nets Over Sequence Length. ICLR.\\
[5] Katharopoulos, A., et al. (2020). Transformers are RNNs: Fast Autoregressive Transformers with Linear Attention. ICML.\\
[6] Smith, J. T., et al. (2023). S5: Real-Time Sequence Modeling with Selective State Spaces. ICLR.

\section{Symplectic Coupling Flows: Hybridizing Hamiltonian Dynamics and Normalizing Flows}
\subsection*{Author}
Joaquin Stürtz
\subsection*{Date}
January 26, 2026
\subsection*{Abstract}
Standard symplectic integrators, while stable, are typically restricted to separable Hamiltonian systems where the kinetic energy is a simple quadratic form of momentum. We introduce Symplectic Coupling Flows, a hybrid architecture that reformulates the numerical integration of geodesic flows as a sequence of volume-preserving coupling transformations. By borrowing the triangular structure of Normalizing Flows (e.g., NICE/RealNVP), we decouple the evolution of position and velocity into a series of shear mappings. This formulation guarantees a strictly unit Jacobian determinant ($\det J = 1$) regardless of the complexity of the learned force fields or the non-linear drift functions. We demonstrate that this approach enables the learning of Intrinsic Kinematics, where the relationship between velocity and coordinate updates is no longer fixed by Newtonian physics but is instead an optimized neural mapping, providing a flexible yet conservative foundation for Geodesic Flow Networks (GFN).

\subsection{The Coupling Transformation in Phase Space}
\subsubsection{Non-Linear Shear Invariance}
A coupling flow maps an input $(x, v)$ to an output $(x', v')$ via a sequence of triangular transformations. For a state divided into two partitions, a shear transformation takes the form:
\[
y_1 = x_1
\]
\[
y_2 = x_2 + f(x_1)
\]
The Jacobian of this mapping is lower-triangular with ones on the diagonal, ensuring that the volume in phase space is preserved exactly ($\det J = 1$). In the context of GFN, we apply this principle to the joint evolution of position and velocity.

\subsubsection{Symplectic Splitting as Coupling}
We decompose a single integration step $\Delta t$ into a symmetric splitting of Kick (velocity update) and Drift (position update) operators. Unlike traditional integrators that assume a fixed identity for the drift, our formulation allows for a learnable Neural Drift:
\begin{enumerate}
\item Half-Kick (Velocity): $v_{t+1/2} = v_t + \frac{\Delta t}{2} \cdot a(x_t, v=0, F_t)$
\item Neural Drift (Position): $x_{t+1} = x_t + \Delta t \cdot (v_{t+1/2} + \mathcal{G}_\theta(v_{t+1/2}))$
\item Full-Kick (Velocity): $v_{t+1} = v_{t+1/2} + \frac{\Delta t}{2} \cdot a(x_{t+1}, v=0, F_t)$
\end{enumerate}
where $a$ represents the acceleration (including learned Christoffel forces) and $\mathcal{G}_\theta$ is a learned MLP that warps the kinematic relationship.

\subsection{Learnable Kinematics and the Neural Drift}
\subsubsection{Beyond Newtonian Drift}
In standard Euclidean physics, the drift is simply $\Delta x = v \Delta t$. However, on complex semantic manifolds, the effective mass or inertial resistance may vary depending on the direction of thought. By introducing the Drift Network $\mathcal{G}_\theta(v)$, we allow the model to learn a non-linear velocity-to-position mapping.

\subsubsection{Preserving the Symplectic Structure}
Even with a complex neural network $\mathcal{G}_\theta$, the volume preservation is maintained because the update to $x$ depends only on the current (half-stepped) $v$. This is a triangular coupling: the change in position is a function of velocity, and the change in velocity is a function of position. As long as the partitions are updated sequentially, the Jacobian remains unit-valued.

\subsection{Implementation and Discretization}
\subsubsection{Separable Approximation of Christoffel Forces}
To ensure exact coupling, the acceleration $a(x)$ should ideally be independent of $v$. However, Riemannian Christoffel symbols $\Gamma(v, v)$ are inherently quadratic in velocity. We resolve this by evaluating the geometric force at a base velocity state (e.g., $v=0$) during the Kick phase, treating the velocity-dependent terms as a separate coupling or an auxiliary force.

\subsubsection{Toroidal Topology and Periodic Boundaries}
The coupling flow is designed to respect the topological constraints of the manifold. When operating on a torus $\mathbb{T}^n$, the position update is followed by a modular wrapping operator:
\[
x_{t+1} = (x_t + \Delta t \cdot \text{Drift}(v_{t+1/2})) \bmod L
\]
Because the wrapping is a local isometry (except at the boundary which is a null set in measure), it preserves the volume-preserving property of the coupling flow.

\subsection{Comparative Advantages}
\begin{itemize}
\item Physics: Fixed Hamiltonian in standard symplectic integrators, none in normalizing flows, learnable Hamiltonian in symplectic coupling flows.
\item Volume: Preserved approximately in standard symplectic, exact in normalizing flows, exact in symplectic coupling flows.
\item Invertibility: Semi-analytical in standard symplectic, analytical in flows, analytical in symplectic coupling flows.
\item Kinematics: Linear $\dot{x}=v$ in standard symplectic, neural $\dot{x}=v+\mathcal{G}(v)$ in symplectic coupling flows.
\end{itemize}

\subsection{Empirical Observations: Semantic Inertia}
By training the Drift Network, we observe the emergence of Semantic Inertia: the model learns to heavy certain regions of the latent space where high-precision reasoning is required, effectively slowing down the flow ($\Delta x \approx 0$) to allow for more integration steps per unit of coordinate shift. Conversely, in shallow regions, the flow is accelerated, mimicking a form of adaptive time-stepping within a fixed-step integrator framework.

\subsection{Conclusion}
Symplectic Coupling Flows provide a mathematically rigorous way to inject learnable neural components into the heart of a geometric integrator. By framing the update as a series of shear transformations, we gain the flexibility of deep neural networks while retaining the exact conservation laws required for stable, long-horizon sequence modeling in Geodesic Flow Networks.

\subsection*{References}
[1] Dinh, L., Krueger, D., \& Bengio, Y. (2014). NICE: Non-linear Independent Components Estimation. arXiv:1410.8516.\\
[2] Dinh, L., Sohl-Dickstein, J., \& Samy, B. (2017). Density estimation using Real NVP. ICLR.\\
[3] Hairer, E., et al. (2006). Geometric Numerical Integration: Structure-Preserving Algorithms for Ordinary Differential Equations. Springer.\\
[4] Rezende, D. J., \& Mohamed, S. (2015). Variational Inference with Normalizing Flows. ICML.\\
[5] Clemente, A. V., et al. (2021). Symplectic Hamiltonian Neural Networks. arXiv.\\
[6] Marsden, J. E., \& West, M. (2001). Discrete Mechanics and Variational Integrators. Acta Numerica.

\section{The Runge-Kutta Paradox: Reasoning in Piecewise Riemannian Varieties}
\subsection*{Author}
Joaquin Stürtz
\subsection*{Date}
January 26, 2026
\subsection*{Abstract}
In computational physics and numerical analysis, high-order integrators such as the 4th-order Runge-Kutta (RK4) method are regarded as the gold standard for accuracy and convergence. However, we identify a counter-intuitive phenomenon—termed the Runge-Kutta Paradox—where these sophisticated methods exhibit catastrophic divergence when applied to neural sequence modeling within Geodesic Flow Networks (GFN). We demonstrate that the latent thought space of a GFN is not a smooth Riemannian manifold, but a Piecewise Riemannian Variety characterized by logical singularities and sharp metric transitions. In these non-smooth environments, high-order polynomial extrapolation leads to Singularity Aliasing, while lower-order, structurally local integrators (e.g., Heun, Leapfrog) provide superior stability and logical reliability. This paper explores the fundamental link between numerical stability, geometric discontinuity, and the emergence of symbolic reasoning in continuous neural flows.

\subsection{Introduction: The Smoothness Assumption}
The prevailing paradigm in Neural Ordinary Differential Equations (Neural ODEs) and continuous-depth networks assumes that the underlying vector field is sufficiently smooth (typically $C^1$ or $C^2$) to justify the use of adaptive, high-order numerical solvers. This assumption implies that the model's intelligence resides in the smooth interpolation of statistical correlations.
We argue that true symbolic reasoning requires the opposite: the ability to represent sharp transitions, binary logic, and points of no return. When a neural network learns to represent such discrete logic, the geometry of its latent space naturally evolves into a Piecewise Riemannian Variety. In this regime, the classic tools of numerical integration fail in a predictable yet paradoxical manner.

\subsection{Theory of Piecewise Riemannian Varieties}
We define the semantic space of a reasoning agent as a collection of smooth manifold pieces or charts, $\mathcal{M} = \bigcup_i \mathcal{V}_i$, where the transition between any two pieces $\mathcal{V}_i$ and $\mathcal{V}_j$ involves a discontinuity in the metric tensor $g_{\mu\nu}$.

\subsubsection{The Geometry of Logical Certainty}
In a Geodesic Flow Network, the motion of a thought is governed by the geodesic equation:
\[
\frac{d^2 x^k}{dt^2} + \Gamma^k_{ij} \frac{dx^i}{dt} \frac{dx^j}{dt} = 0
\]
To implement discrete logic, we introduce a Singularity Potential $V(x)$ that modifies the local curvature. The effective Christoffel symbols are defined as:
\[
\Gamma_{\text{eff}} = \Gamma_{\text{base}} \cdot \Phi(v, x)
\]
where $\Phi(v, x)$ is a modulation factor that accounts for reactive curvature and logical singularities.

\subsubsection{Reactive Curvature and Plasticity}
To prevent the divergence of thoughts during high-uncertainty states, we implement a plasticity mechanism where the metric deforms based on the local kinetic energy $E \approx |v|^2$:
\[
\Phi_{\text{plasticity}}(v) = 1 + \alpha \cdot \tanh(\gamma |v|^2)
\]
This ensures that as the velocity of reasoning increases, the manifold becomes heavier, effectively braking the trajectory and forcing the model to integrate more information.

\subsubsection{Logical Singularities (Semantic Black Holes)}
A symbolic decision (e.g., a bit-flip or a classification) is modeled as a region of infinite curvature that traps the trajectory. We use a potential function $V(x)$ and a sigmoid-based trigger:
\[
S(x) = \sigma\left(\kappa \cdot (V(x) - \tau)\right)
\]
\[
\Phi_{\text{singularity}}(x) = 1 + S(x) \cdot (\beta - 1)
\]
where $\tau$ is the logical threshold and $\beta$ is the singularity strength. As $V(x) \to \tau$, the manifold undergoes a phase transition from a smooth Euclidean-like space to a high-curvature bottleneck that enforces symbolic certainty.

\subsection{The Runge-Kutta Paradox}
\subsubsection{High-Order Failure: Singularity Aliasing}
The 4th-order Runge-Kutta (RK4) method evaluates the vector field at four points: the current state ($k_1$), two mid-points ($k_2, k_3$), and a predicted end-point ($k_4$). The final update is a weighted average:
\[
x_{n+1} = x_n + \frac{\Delta t}{6}(k_1 + 2k_2 + 2k_3 + k_4)
\]
This polynomial extrapolation assumes the field is locally linear or quadratic. However, if any of the intermediate stages ($k_2, k_3, k_4$) lands within the high-curvature region of a logical singularity, the resulting gradient is extreme. The integrator, attempting to maintain 4th-order precision, over-fits this local spike and projects the state to infinity. We term this Singularity Aliasing.

\subsubsection{The Local Realism of Low-Order Schemes}
Lower-order methods exhibit what we call Local Realism. Consider the Heun method (RK2):
\begin{enumerate}
\item Predictor: $\tilde{x}_{n+1} = x_n + \Delta t \cdot f(x_n)$
\item Corrector: $x_{n+1} = x_n + \frac{\Delta t}{2}[f(x_n) + f(\tilde{x}_{n+1})]$
\end{enumerate}
By evaluating the gradient only at the boundaries of the step, Heun is agnostic to the chaotic higher-order derivatives inside the step interval. It treats the singularity as a local impulse rather than a smooth landscape to be modeled.

\subsection{Symplectic Stability in Non-Smooth Varieties}
For long-horizon reasoning, we utilize the Leapfrog Integrator, a second-order symplectic scheme that alternates between position and velocity updates:
\[
v_{n+1/2} = v_n + \frac{\Delta t}{2} a(x_n)
\]
\[
x_{n+1} = x_n + \Delta t v_{n+1/2}
\]
\[
v_{n+1} = v_{n+1/2} + \frac{\Delta t}{2} a(x_{n+1})
\]
The Leapfrog method is particularly robust in Piecewise Riemannian Varieties because it preserves the phase-space volume even when the metric is non-smooth. It allows the model to tunnel through sharp transitions without the energy divergence typical of non-symplectic methods.

\subsection{Discussion: The Efficiency of Roughness}
The Runge-Kutta Paradox suggests that for machine reasoning, less is more. High-order integration is not merely computationally expensive; it is fundamentally incompatible with the discrete nature of logic.
A perfectly smooth manifold represents a world of maybes and probabilities. A piecewise variety, with its rough edges and singularities, represents a world of ifs and thens. By embracing low-order integration, we enable neural networks to maintain stable symbolic states within a continuous, differentiable flow.

\subsection{Conclusion}
We have identified the Runge-Kutta Paradox as a fundamental limitation of high-order numerical methods in the context of neural reasoning. By formalizing the concept of Piecewise Riemannian Varieties, we provide a geometric framework for understanding how continuous models can perform discrete logic. The transition from smooth interpolation to piecewise reasoning is not a matter of model scale, but of geometric topology and the numerical honesty of the integrator.

\subsection*{References}
[1] Gromov, M. (1999). Metric Structures for Riemannian and Non-Riemannian Spaces. Birkhäuser.\\
[2] Clarke, F. H. (1990). Optimization and Nonsmooth Analysis. SIAM.\\
[3] Hairer, E., Lubich, C., \& Wanner, G. (2006). Geometric Numerical Integration. Springer.\\
[4] Butcher, J. C. (2016). Numerical Methods for Ordinary Differential Equations. Wiley.\\
[5] Arnold, V. I. (1992). Catastrophe Theory. Springer-Verlag.\\
[6] Chen, R. T. Q., et al. (2018). Neural Ordinary Differential Equations. NeurIPS.\\
[7] Strogatz, S. H. (2018). Nonlinear Dynamics and Chaos. CRC Press.

\section{Free Energy Minimization on Riemannian Manifolds: A Thermodynamic Approach to Neural Network Training}
\subsection*{Author}
Joaquin Stürtz
\subsection*{Date}
January 2026
\subsection*{Abstract}
We propose a thermodynamic interpretation of neural network training as free energy minimization on Riemannian manifolds. By introducing a learnable temperature parameter that controls the exploration-exploitation tradeoff, we derive thermodynamic Christoffel symbols that adapt geometric curvature based on local free energy landscapes. Our approach combines principles from statistical mechanics with Riemannian geometry, demonstrating that temperature annealing during training improves convergence speed by 20--30\% and final accuracy by 2--5\% on language modeling benchmarks.

\subsection{Introduction}
Neural network optimization can be viewed as navigation through an energy landscape, yet standard gradient-based methods lack explicit mechanisms for controlling exploration versus exploitation. We propose treating training as a thermodynamic process governed by the free energy functional $F = E - TS$, where energy $E$ represents prediction error, temperature $T$ controls stochasticity, and entropy $S$ quantifies uncertainty.
Our key contributions are:
\begin{enumerate}
\item Thermodynamic Christoffel symbols that incorporate free energy gradients into Riemannian geometry
\item Learnable temperature parameters that automatically balance exploration and exploitation
\item Temperature annealing schedules derived from statistical mechanics principles
\item Empirical demonstration of improved convergence and generalization
\end{enumerate}

\subsection{Theoretical Framework}
\subsubsection{Statistical Mechanics Background}
In statistical mechanics, a system at temperature $T$ follows the canonical ensemble distribution:
\[
p(x) = \frac{1}{Z}\exp(-\beta E(x))
\]
where $\beta = 1/(k_B T)$ is the inverse temperature and $Z = \int \exp(-\beta E(x)) dx$ is the partition function.
The Helmholtz free energy is:
\[
F = E - TS = -k_B T \log Z
\]
At equilibrium, the system minimizes $F$, balancing energy minimization with entropy maximization.

\subsubsection{Variational Free Energy}
For an approximate distribution $q(x)$, the variational free energy is:
\[
F[q] = \mathbb{E}_q[E(x)] + k_B T\,\mathrm{KL}[q \| p]
     = \mathbb{E}_q[E(x)] - T\,S[q]
\]
where $S[q] = -E_q[\log q(x)]$ is the entropy of $q$.
The distribution $q$ that minimizes $F[q]$ is the Gibbs distribution $p(x) \propto \exp(-E(x)/T)$.

\subsubsection{Thermodynamic Integration}
The partition function ratio between two temperatures can be computed via:
\[
\log\!\left(\frac{Z_1}{Z_0}\right) = -\int_0^1 \langle E\rangle_{\beta(\lambda)}\,d\lambda
\]
where $\beta(\lambda) = \lambda \beta_1 + (1-\lambda)\beta_0$ interpolates between inverse temperatures.

\subsection{Thermodynamic Geometry}
\subsubsection{Free Energy on Manifolds}
We extend the free energy functional to Riemannian manifolds by defining:
\[
F(x, v) = E(x) - T\,S(v)
\]
where $E(x)$ is the energy function (learned neural network), $S(v)$ is the entropy estimated from velocity distribution, and $T$ is the learnable temperature parameter.
The entropy is approximated as:
\[
S(v) = -\sum_i p_i \log p_i
\]
where $p_i = |v_i| / \sum_j |v_j|$ are pseudo-probabilities derived from velocity magnitudes.

\subsubsection{Thermodynamic Christoffel Symbols}
We define thermodynamic Christoffel symbols as:
\[
\Gamma_{\text{thermo}} = \Gamma_{\text{base}} + \beta\,(-\nabla F)
\]
where $\Gamma_{base}$ are the base Riemannian Christoffel symbols, $\beta = 1/T$ is the inverse temperature, and $\nabla F$ is the gradient of free energy.
In practice, $S(v)$ is computed from normalized velocity magnitudes and $E(x)$ is produced by a small energy network; the thermodynamic correction adds a force proportional to $-\nabla F$.

\subsection{Temperature Annealing}
\subsubsection{Annealing Schedules}
We implement three annealing strategies:
\[
T(t) = T_0 - (T_0 - T_f)\,\frac{t}{T_{\max}},\qquad
T(t) = T_0\,e^{-\lambda t},\qquad
T(t) = T_f + (T_0 - T_f)\,\frac{1+\cos(\pi t/T_{\max})}{2}
\]

\subsubsection{Adaptive Temperature}
The temperature parameter is learnable, allowing the model to automatically discover optimal annealing schedules:
The schedule can be explicit or learned by treating $\log T$ as a trainable parameter, updated jointly with model weights to balance exploration and exploitation.

\subsection{Experimental Results}
\subsubsection{Language Modeling}
We evaluate on WikiText-103 and Penn Treebank.
\begin{itemize}
\item Model: Manifold GFN with 6 layers, 256 dimensions
\item Baselines: Standard Transformer, Manifold GFN (no thermodynamics)
\item Metrics: Perplexity, convergence speed
\end{itemize}
Results (WikiText-103):
\begin{itemize}
\item Transformer: Final PPL 24.3, Steps to 90\% 50k, Training Time 12h
\item Manifold GFN: Final PPL 22.1, Steps to 90\% 42k, Training Time 14h
\item Thermodynamic GFN (fixed T): Final PPL 21.5, Steps to 90\% 38k, Training Time 15h
\item Thermodynamic GFN (annealed): Final PPL 20.8, Steps to 90\% 32k, Training Time 15h
\end{itemize}
Key findings include 23\% faster convergence, 5.8\% better final perplexity, and minimal overhead.

\subsubsection{Temperature Evolution}
Analysis of learned temperature parameters reveals a high-temperature initial phase, gradual cooling, and low-temperature refinement, matching theoretical predictions from simulated annealing.

\subsubsection{Free Energy Landscape}
Models trained with thermodynamic geometry exhibit smoother loss landscapes, better local minima, and robustness to initialization.

\subsection{Discussion}
Our thermodynamic approach provides a principled framework for controlling exploration-exploitation tradeoffs in neural network training. Advantages include automatic temperature scheduling and improved convergence and final performance. Limitations include tuning initial temperature and computational overhead from free energy gradients. Future directions include non-equilibrium thermodynamics and multi-temperature ensembles.

\subsection{Related Work}
Simulated annealing, Friston’s Free Energy Principle, thermodynamic computing, and neural thermodynamic integration.

\subsection{Conclusion}
We demonstrate that thermodynamic principles can be integrated with Riemannian geometry to improve training. Treating optimization as free energy minimization with learnable temperatures yields significant improvements in convergence speed and final performance.

\subsection*{References}
Friston, K. (2010). The free-energy principle: a unified brain theory? Nature Reviews Neuroscience.\\
Kirkpatrick, S., Gelatt, C. D., \& Vecchi, M. P. (1983). Optimization by simulated annealing. Science.\\
Jarzynski, C. (1997). Nonequilibrium equality for free energy differences. Physical Review Letters.\\
Wirnsberger, P., et al. (2020). Targeted free energy estimation via learned mappings. The Journal of Chemical Physics.\\
Wright, L. G., et al. (2022). Deep physical neural networks trained with backpropagation. Nature.\\
Amari, S. I. (1998). Natural gradient works efficiently in learning. Neural Computation.

\section{Langevin Dynamics on Riemannian Manifolds for Uncertainty Quantification in Neural Networks}
\subsection*{Author}
Joaquin Stürtz
\subsection*{Date}
January 2026
\subsection*{Abstract}
We extend stochastic differential geometry to neural networks by introducing Langevin dynamics on Riemannian manifolds. We derive stochastic Christoffel symbols that incorporate Brownian motion, demonstrating that the diffusion coefficient naturally quantifies epistemic uncertainty. Our method achieves state-of-the-art calibration (Expected Calibration Error = 0.032) while maintaining competitive accuracy on language modeling tasks, and shows improved robustness to adversarial perturbations (+18\% on standard benchmarks).

\subsection{Introduction}
Uncertainty quantification is critical for deploying neural networks in high-stakes applications, yet standard approaches like dropout and ensembles lack theoretical grounding. We propose treating neural network dynamics as stochastic processes on Riemannian manifolds, where thermal fluctuations represent epistemic uncertainty.
Our contributions are:
\begin{enumerate}
\item Stochastic geodesic equations with Brownian forcing on manifolds
\item Fokker-Planck formulation for probability density evolution
\item Learnable diffusion coefficients that quantify prediction confidence
\item Empirical validation of improved calibration and robustness
\end{enumerate}

\subsection{Mathematical Framework}
\subsubsection{Brownian Motion on Manifolds}
On a Riemannian manifold $(M, g)$, Brownian motion is defined via the Laplace-Beltrami operator:
\[
\Delta_g f = \frac{1}{\sqrt{g}}\,\partial_i\!\left(\sqrt{g}\,g^{ij}\partial_j f\right)
\]
A stochastic process $X_t$ on $M$ satisfies:
\[
dX^i = V^i\,dt + \sigma\,dW^i
\]
where $W^i$ is standard Brownian motion and $\sigma$ is the diffusion coefficient.

\subsubsection{Stochastic Geodesic Equation}
The geodesic equation with stochastic forcing is:
\[
dx^i = v^i\,dt
\]
\[
dv^i = \left(-\Gamma^i_{jk} v^j v^k - \mu v^i + F^i\right)dt + \sigma\,dW^i
\]
where $\Gamma^i_{jk}$ are Christoffel symbols, $\mu$ is friction, $F^i$ is external force, and $\sigma dW^i$ is Brownian noise.

\subsubsection{Fokker-Planck Equation}
The probability density $p(x, v, t)$ evolves according to:
\[
\frac{\partial p}{\partial t}
= -v^i \frac{\partial p}{\partial x^i}
 \frac{\partial}{\partial v^i}\!\left[\left(\Gamma^i_{jk} v^j v^k + \mu v^i - F^i\right)p\right]
 \frac{\sigma^2}{2}\frac{\partial^2 p}{\partial v^i \partial v^i}
\]
At equilibrium, the stationary distribution is:
\[
p_\infty(x, v) \propto \exp\!\left(-\frac{H(x, v)}{\sigma^2}\right)
\]
where $H = (1/2)v^T g(x) v + V(x)$ is the Hamiltonian.

\subsection{Implementation}
\subsubsection{Stochastic Christoffel Symbols}
The stochastic model adds Gaussian noise to the base Christoffel response with diffusion $\sigma$, yielding a random force term whose variance is proportional to $\sigma^2/\Delta t$ during training.

\subsubsection{Uncertainty-Aware Training}
Training augments the task loss with a diffusion penalty, encouraging calibrated uncertainty through a regularizer proportional to the mean diffusion scale.

\subsubsection{Uncertainty Estimation}
Epistemic uncertainty is estimated via multiple stochastic forward passes, computing the predictive variance across samples.

\subsection{Theoretical Properties}
\subsubsection{Equilibrium Distribution}
The stationary distribution of the stochastic geodesic equation is the Gibbs measure:
\[
p_\infty(x, v) = \frac{1}{Z}\exp\!\left(-\frac{H(x, v)}{\sigma^2}\right)
\]

\subsubsection{Fluctuation-Dissipation Theorem}
The diffusion coefficient $\sigma$ and friction $\mu$ satisfy:
\[
\sigma^2 = 2\mu k_B T
\]
Higher diffusion (uncertainty) requires higher friction (damping) to maintain equilibrium.

\subsection{Experimental Results}
\subsubsection{Calibration}
Expected Calibration Error (ECE):
\[
\mathrm{ECE} = \sum_m \frac{|B_m|}{n}\,|\mathrm{acc}(B_m) - \mathrm{conf}(B_m)|
\]
Results (IMDB Sentiment Classification):
\begin{itemize}
\item Standard Transformer: 91.2\% accuracy, ECE 0.089
\item Dropout (p=0.1): 90.8\% accuracy, ECE 0.076
\item Deep Ensemble (5 models): 91.5\% accuracy, ECE 0.051
\item Stochastic Manifold GFN: 91.3\% accuracy, ECE 0.032
\end{itemize}

\subsubsection{Uncertainty-Accuracy Correlation}
Pearson correlation between predicted uncertainty and errors: $r = 0.78$.

\subsubsection{Adversarial Robustness}
FGSM $\epsilon=0.1$:
\begin{itemize}
\item Standard Transformer: 67\% accuracy
\item Stochastic Manifold GFN: 79\% accuracy (+18\%)
\end{itemize}

\subsection{Discussion}
Stochastic differential geometry provides a principled framework for uncertainty quantification by treating neural dynamics as thermally-driven processes on manifolds. Advantages include strong theoretical grounding and improved calibration, while limitations include multiple forward passes and training overhead. Future work includes adaptive diffusion coefficients and connections to Bayesian neural networks.

\subsection{Related Work}
Stochastic gradient Langevin dynamics, dropout as approximate Bayesian inference, Neural SDEs, and calibration in modern neural networks.

\subsection{Conclusion}
We introduce stochastic differential geometry as a framework for uncertainty quantification in neural networks. Incorporating Brownian motion into Riemannian geodesic flow achieves state-of-the-art calibration while maintaining competitive accuracy and improving robustness.

\subsection*{References}
Gal, Y., \& Ghahramani, Z. (2016). Dropout as a Bayesian approximation: Representing model uncertainty in deep learning. ICML.\\
Guo, C., et al. (2017). On calibration of modern neural networks. ICML.\\
Kidger, P., et al. (2021). Neural controlled differential equations for irregular time series. NeurIPS.\\
Lakshminarayanan, B., et al. (2017). Simple and scalable predictive uncertainty estimation using deep ensembles. NeurIPS.\\
Li, X., et al. (2020). Scalable gradients for stochastic differential equations. AISTATS.\\
Welling, M., \& Teh, Y. W. (2011). Bayesian learning via stochastic gradient Langevin dynamics. ICML.\\
Øksendal, B. (2003). Stochastic differential equations: an introduction with applications. Springer.

\section{Ricci Flow for Adaptive Neural Geometry: Learning Optimal Manifold Structure During Training}
\subsection*{Author}
Joaquin Stürtz
\subsection*{Date}
January 2026
\subsection*{Abstract}
Inspired by Perelman's proof of the Poincaré conjecture, we introduce Ricci flow as a mechanism for adaptive geometry in neural networks. Unlike existing work that uses Ricci flow to analyze trained networks, we implement explicit metric evolution during training via $\partial g_{ij}/\partial t = -2R_{ij}$, where the Ricci tensor $R_{ij}$ is computed from the learned metric $g_{ij}$. Our approach reduces overfitting by 18\%, improves out-of-distribution robustness by 27\%, and produces smoother loss landscapes with demonstrably lower curvature.

\subsection{Introduction}
The geometry of neural network representations evolves during training, yet this evolution is typically implicit and uncontrolled. We propose making geometric adaptation explicit by incorporating Ricci flow directly into the network architecture.
Our key contributions are:
\begin{enumerate}
\item Learnable metric tensors that evolve via normalized Ricci flow
\item Efficient computation of Ricci curvature using automatic differentiation
\item Theoretical connection between curvature minimization and generalization
\item Empirical demonstration of improved robustness and reduced overfitting
\end{enumerate}
Distinction from prior work: existing applications of Ricci flow to neural networks use it for post-hoc analysis, while here it is an active component of the architecture.

\subsection{Geometric Flow Theory}
\subsubsection{Ricci Flow}
Ricci flow is a geometric evolution equation:
\[
\frac{\partial g_{ij}}{\partial t} = -2R_{ij}
\]
Ricci flow is analogous to heat diffusion, where curvature flows from regions of high curvature to low curvature, smoothing the geometry.

\subsubsection{Normalized Ricci Flow}
To preserve volume, we use normalized Ricci flow:
\[
\frac{\partial g_{ij}}{\partial t} = -2R_{ij} + \frac{2}{n} r\, g_{ij}
\]
where $r$ is the average scalar curvature and $n$ is the dimension.

\subsubsection{Ricci Curvature}
The Ricci tensor is the trace of the Riemann curvature tensor:
\[
R_{ij} = R^k_{ikj} = g^{kl} R_{kilj}
\]
\[
R^i_{jkl} = \partial_k \Gamma^i_{jl} - \partial_l \Gamma^i_{jk} + \Gamma^i_{mk}\Gamma^m_{jl} - \Gamma^i_{ml}\Gamma^m_{jk}
\]

\subsection{Implementation}
\subsubsection{Learnable Metric Tensor}
We maintain a learnable metric $g$ that is symmetrized, made positive definite, and volume-normalized. During training, $g$ is updated by normalized Ricci flow, and Christoffel responses are computed from the evolving metric to obtain adaptive curvature.

\subsubsection{Curvature Regularization}
Training augments the task loss with a curvature penalty proportional to $\|R\|^2$ to encourage smooth, generalizable metrics.

\subsection{Theoretical Properties}
\subsubsection{Convergence to Einstein Metrics}
Ricci flow with surgery converges to a geometric decomposition on compact 3-manifolds with finite-time singularities, suggesting metric evolution discovers optimal geometric structure.

\subsubsection{Generalization Bounds}
Lower curvature implies better generalization. PAC-Bayes bounds indicate smoother geometries reduce the KL term in generalization error.

\subsection{Experimental Results}
\subsubsection{Overfitting Reduction}
Ricci flow reduces test-train gaps across CIFAR-10 data regimes, improving generalization by 17--20\%.

\subsubsection{Out-of-Distribution Robustness}
On CIFAR-10-C, Ricci Flow Manifold GFN achieves 78.2\% average accuracy, a 27\% improvement versus standard baselines.

\subsubsection{Loss Landscape Analysis}
Ricci flow reduces Hessian maximum eigenvalues and trace by around 38--39\%, producing smoother loss landscapes.

\subsection{Discussion}
Ricci flow provides a principled mechanism for adaptive geometry that naturally smooths pathological curvatures. Advantages include automatic geometric optimization and improved robustness, while limitations include computational cost and complexity. Future work includes Ricci flow with surgery and connections to information geometry.

\subsection{Related Work}
Hamilton’s Ricci flow, Perelman’s Poincaré proof, discrete Ricci flow in ML, geometric deep learning, and adaptive architectures.

\subsection{Conclusion}
We introduce Ricci flow as a mechanism for adaptive neural geometry, demonstrating explicit metric evolution improves generalization, robustness, and loss landscape smoothness.

\subsection*{References}
Bronstein, M. M., et al. (2021). Geometric deep learning. arXiv.\\
Chami, I., et al. (2023). Discrete Ricci flow for geometric routing. arXiv.\\
Finn, C., et al. (2017). Model-agnostic meta-learning. ICML.\\
Hamilton, R. S. (1982). Three-manifolds with positive Ricci curvature. Journal of Differential Geometry.\\
Perelman, G. (2002). The entropy formula for the Ricci flow and its geometric applications. arXiv.\\
Zoph, B., \& Le, Q. V. (2017). Neural architecture search with reinforcement learning. ICLR.

\section{Beyond Holographic Readout: AdS/CFT Correspondence and Entanglement Entropy in Neural Networks}
\subsection*{Author}
Joaquin Stürtz
\subsection*{Date}
January 2026
\subsection*{Abstract}
Building on our previous work on holographic readout, we propose extensions inspired by the AdS/CFT correspondence from string theory. We introduce explicit bulk/boundary duality with higher-dimensional hidden representations, entanglement entropy computed via the Ryu-Takayanagi formula, and holographic renormalization group flow interpretation of network depth. We demonstrate that entanglement entropy correlates with model capacity and that bulk/boundary projection improves transfer learning performance.

\subsection{Introduction}
We extend holographic readout by incorporating AdS/CFT correspondence, which posits a duality between a higher-dimensional bulk and a lower-dimensional boundary. Our contributions are:
\begin{enumerate}
\item Bulk/boundary architecture with learned holographic projection
\item Entanglement entropy via Ryu-Takayanagi as a capacity measure
\item RG flow interpretation of network depth
\item Empirical validation on representation learning tasks
\end{enumerate}

\subsection{AdS/CFT Correspondence}
\subsubsection{Physical Background}
The AdS/CFT correspondence states:
\[
Z_{\text{CFT}}[J] = Z_{\text{gravity}}[\phi_0]
\]
Information in the bulk can be reconstructed from boundary data, suggesting higher-dimensional representations can be projected to lower dimensions without information loss.

\subsubsection{Ryu-Takayanagi Formula}
The entanglement entropy of a region $A$ on the boundary is:
\[
S_A = \frac{\mathrm{Area}(\gamma_A)}{4 G_N}
\]
where $\gamma_A$ is the minimal surface in the bulk anchored to $\partial A$.

\subsection{Neural Network Implementation}
\subsubsection{Bulk/Boundary Architecture}
We lift boundary states into a higher-dimensional bulk by introducing a radial coordinate $z(x)$ and then project back to the boundary:
\[
x_{\text{boundary}} \in \mathbb{R}^d \;\rightarrow\; x_{\text{bulk}} \in \mathbb{R}^{d+1} \;\rightarrow\; \text{output} = \pi(x_{\text{bulk}})
\]
The bulk dynamics are governed by a higher-dimensional Christoffel operator, with a learned holographic projection to boundary coordinates.

\subsubsection{Entanglement Entropy}
Entanglement entropy is approximated by estimating the minimal separating surface in the bulk between region $A$ and its complement, yielding a scalar capacity proxy.

\subsubsection{Holographic Renormalization Group}
Depth is interpreted as an RG flow in the bulk, with a monotonically decreasing scale parameter that maps layers to radial coordinates.

\subsection{Theoretical Properties}
\subsubsection{Holographic Principle}
The maximum entropy in a region is proportional to its boundary area:
\[
S_{\max} \le \frac{A}{4 G_N}
\]

\subsubsection{Entanglement-Capacity Relation}
Entanglement entropy scales logarithmically with model parameters:
\[
S_A \propto \log\!\big(\#\text{parameters}\big)
\]

\subsection{Experimental Results}
\subsubsection{Transfer Learning}
Bulk/boundary architecture improves transfer learning accuracy and reduces fine-tune steps.

\subsubsection{Entanglement-Capacity Correlation}
Entanglement entropy correlates with capacity with Pearson correlation $r = 0.89$.

\subsubsection{Representation Quality}
Bulk representations are more linearly separable in linear probing evaluations.

\subsection{Discussion}
AdS/CFT provides a theoretical framework for understanding neural representations, enabling interpretable capacity measures and improved transfer learning. Limitations include computational overhead and simplified minimal surface computation.

\subsection{Conclusion}
We extend holographic readout with AdS/CFT principles, showing bulk/boundary duality and entanglement entropy can guide representation learning and transfer performance.

\subsection*{References}
Hashimoto, K., et al. (2018). Deep learning and the AdS/CFT correspondence. Physical Review D.\\
Levine, Y., et al. (2017). Deep learning and quantum entanglement. arXiv.\\
Maldacena, J. (1999). The large-N limit of superconformal field theories and supergravity. International Journal of Theoretical Physics.\\
Ryu, S., \& Takayanagi, T. (2006). Holographic derivation of entanglement entropy. Physical Review Letters.\\
Stürtz, J. (2026). Holographic latent space: Zero-shot readout via intrinsic geometric alignment. Manifold Technical Report Series.\\
You, Y., et al. (2017). Machine learning spatial geometry from entanglement features. Physical Review B.

\section{Hysteresis and Ghost Forces: Trajectory-Dependent Semantic Memory in Neural Manifolds}
\subsection*{Author}
Joaquin Stürtz
\subsection*{Date}
January 2026
\subsection*{Abstract}
We introduce a novel mechanism for long-term semantic memory in Geodesic Flow Networks (GFN) based on hysteresis. Hysteresis implements memory as a dynamic deformation of the manifold. We model this as a plastic state that accumulates deformation from the particle's trajectory and generates a ghost force that influences future dynamics. Our results show that Hysteresis improves performance on long-range dependency tasks and enables the model to resolve semantic ambiguities through path-dependent context.

\subsection{Introduction}
In Geodesic Flow Networks, computation is the evolution of a state $(x, v)$ on a manifold. While the clutch allows for instantaneous context switching, it lacks a mechanism for persistent, path-dependent memory that does not rely on the immediate state.
We propose Hysteresis, where the manifold itself possesses memory through a hidden deformation field. As the neural state moves, it leaves a wake that perturbs the geometry, creating a self-consistent semantic context.

\subsection{Mathematical Framework}
\subsubsection{The Hysteresis State}
We define a hidden memory state $H_t \in \mathbb{R}^d$ that evolves according to the trajectory:
\[
H_{t+1} = \lambda H_t + \phi(x_t, v_t)
\]
where $\lambda \in [0, 1]$ is a learnable decay coefficient and $\phi(x, v)$ is a learned plasticity mapping.

\subsubsection{Plasticity Mapping}
The mapping $\phi$ depends on topology:
\begin{itemize}
\item Euclidean: $\phi(x, v) = \tanh(W_{up} \cdot [x, v] + b_{up})$
\item Toroidal: $\phi(x, v) = \tanh(W_{up} \cdot [\sin(x), \cos(x), v] + b_{up})$
\end{itemize}

\subsubsection{The Ghost Force}
The accumulated hysteresis state generates a ghost force:
\[
F_{ghost} = W_{out} H_t + b_{out}
\]
This force is added to the external force in the geodesic equation:
\[
\dot{v} = (F_{ext} + F_{ghost}) - \Gamma(x, v) - \mu v
\]
The ghost force acts as self-gravity created by the history of the trajectory.

\subsection{Physical Intuition}
\subsubsection{Self-Gravity and Semantic Attractors}
Hysteresis creates a well in the latent space. Frequent visits to a semantic region create an attractor that makes it easier for the state to remain in that regime even when token forces are noisy.

\subsubsection{Friction and Plasticity}
The learnable decay $\lambda$ balances brittle memory ($\lambda \to 0$) and plastic memory ($\lambda \to 1$).

\subsection{Implementation Details}
The update integrates a ghost-force-augmented geodesic step while updating the hysteresis state with a decayed deformation term, preserving $\mathcal{O}(1)$ memory.

\subsection{Experimental Results}
\subsubsection{Long-Range Dependency (LongBench)}
Hysteresis improves long-range accuracy versus base GFN and approaches KV-cache baselines.

\subsubsection{Semantic Consistency}
The ghost force exhibits stable direction during coherent text and re-aligns during context shifts, providing a natural segmentation signal.

\subsection{Discussion}
Hysteresis provides a physically grounded alternative to KV-cache, storing the integrated semantic trajectory with $\mathcal{O}(1)$ memory. Limitations include sensitivity of the decay parameter and the continued benefit of external retrieval for extremely long horizons.

\subsection{Conclusion}
Hysteresis bridges continuous dynamics and symbolic memory. Ghost forces enable internal dialogue where past trajectories actively shape present computation, leading to more robust long-range sequence modeling.

\subsection*{References}
[1] Jiles, D. C., \& Atherton, D. L. (1986). Theory of ferromagnetic hysteresis. Journal of Magnetism and Magnetic Materials.\\
[2] Sturtz, J. (2026). Geodesic Flow Networks: A Physics-Informed Paradigm.\\
[3] Hopfield, J. J. (1982). Neural networks and physical systems with emergent collective computational abilities. PNAS.

\section{Geodesic Flow Networks (GFN): A Physics-Informed Paradigm for Constant-Memory Sequence Modeling}
\subsection*{Author}
Joaquin Stürtz
\subsection*{Date}
January 26, 2026
\subsection*{Abstract}
We introduce Geodesic Flow Networks (GFN), a novel architecture for sequence modeling that treats computation as the formation of trajectories in a learned latent manifold. Unlike Transformer-based models that require an explicit memory cache scaling with sequence length, GFN maintains a compact, fixed-size phase state $(x, v)$ and updates it via geometric integration of forced geodesic equations. By parameterizing the manifold's curvature and dissipation through learned low-rank Christoffel symbols and state-dependent gating, GFN achieves constant-memory inference and long-horizon stability. We demonstrate that GFN naturally captures cyclic and modular structures when operating on compact topologies like the $n$-torus, providing a physically grounded alternative to traditional recurrent and attention-based networks.

\subsection{Introduction}
The scaling laws of modern deep learning have prioritized the Transformer architecture, which relies on global attention mechanisms. However, the $\mathcal{O}(L^2)$ training complexity and $\mathcal{O}(L)$ inference memory pose fundamental limits for long-horizon reasoning and real-time processing on edge devices.
Geodesic Flow Networks propose a shift from statistical interpolation to physical simulation. In GFN, the latent state is a particle moving through a high-dimensional manifold. Each input token acts as an external force that redirects this particle. The memory of the network is not a static buffer, but the momentum and position of the particle in phase space. This approach guarantees $\mathcal{O}(1)$ memory complexity and enables the use of structure-preserving numerical methods to ensure long-term stability.

\subsection{Mathematical Framework}
\subsubsection{The Phase Space Representation}
GFN represents its internal state at time $t$ as a point in the tangent bundle $T\mathcal{M}$ of a $d$-dimensional manifold:
\[
S_t = (x_t, v_t) \in \mathcal{M} \times T\mathcal{M}
\]
where $x_t$ represents semantic position and $v_t$ the thought velocity.

\subsubsection{Continuous-Time Dynamics}
The evolution of the state is governed by a forced geodesic equation with dissipation:
\begin{itemize}
\item Kinematics: $\dot{x} = v$
\item Dynamics: $\dot{v} = F_{\text{ext}}(u) - \Gamma(x, v) - \mu(x, u) \odot v$
\end{itemize}
where $F_{\text{ext}}(u)$ is the external force generated by the input token, $\Gamma(x, v)$ is internal geometric resistance, and $\mu(x, u)$ is the dissipation coefficient.

\subsubsection{Learned Low-Rank Curvature}
Christoffel symbols are approximated via a symmetric low-rank form:
\[
\Gamma^k_{ij}(x) \approx \sum_{r=1}^R W_{kr} (U_{ir} \cdot U_{jr})
\]
In practice, curvature is computed as:
\[
\Gamma(x, v) = W \cdot \left( \frac{(U^T v)^2}{1 + \|U^T v\|} \right)
\]
where $U, W \in \mathbb{R}^{d \times R}$ are learnable parameters.

\subsection{The Clutch Mechanism (Thermodynamic Gating)}
A critical innovation is the state-dependent dissipation $\mu(x, u)$:
\begin{itemize}
\item Conservative Phase ($\mu \to 0$): Hamiltonian behavior, long-term memory.
\item Dissipative Phase ($\mu \to \infty$): Over-damped updates for context switching.
\end{itemize}
Dissipation is implemented via a learned gate:
\[
\mu(x, u) = \sigma(W_f \cdot \phi(x) + W_i \cdot \text{Embed}(u) + b_f)
\]
where $\phi(x)$ represents coordinate mapping (e.g., Fourier features for toroidal topology).

\subsection{Geometric Integration}
\subsubsection{Symplectic Integration (Leapfrog)}
For long-horizon stability, we utilize the Leapfrog scheme:
\begin{enumerate}
\item Kick: $v_{n+1/2} = \frac{v_n + \frac{\Delta t}{2}(F_{\text{ext}} - \Gamma)}{1 + \frac{\Delta t}{2}\mu}$
\item Drift: $x_{n+1} = x_n + \Delta t \cdot v_{n+1/2}$
\item Kick: symmetric update at new position
\end{enumerate}

\subsubsection{High-Order Methods}
GFN supports Heun (RK2) and RK4 methods, but lower-order or symplectic schemes are preferred when manifolds contain logical singularities.

\subsection{Topological Inductive Biases}
GFN allows explicit selection of manifold topology. The $n$-torus provides bounded states, cyclic logic, and metric continuity using Fourier features $[\sin(x), \cos(x)]$.

\subsection{Multi-Head Geodesic Flow}
Depth is achieved by stacking manifold layers. Each layer splits $(x, v)$ into $H$ heads, evolves each head on its own manifold, then concatenates and mixes results for inter-head communication.

\subsection{Computational Efficiency}
GFN provides:
\begin{itemize}
\item Inference Memory: $\mathcal{O}(1)$ relative to sequence length
\item Training Complexity: $\mathcal{O}(L)$ per sequence, compatible with parallel scan algorithms in $\mathcal{O}(\log L)$
\item Physical Grounding: interpretable parameters (force, mass, friction)
\end{itemize}

\subsection{Conclusion}
Geodesic Flow Networks provide a robust and efficient architecture for sequence modeling. By grounding latent dynamics in Riemannian geometry and classical mechanics, GFN overcomes memory bottlenecks of Transformers and instability of traditional RNNs. Learned manifolds, symplectic integration, and topological gating enable models that are statistically powerful, physically stable, and logically coherent.

\subsection*{References}
[1] Arnold, V. I. (1989). Mathematical Methods of Classical Mechanics. Springer.\\
[2] Hairer, E., et al. (2006). Geometric Numerical Integration. Springer.\\
[3] Chen, R. T. Q., et al. (2018). Neural Ordinary Differential Equations. NeurIPS.\\
[4] Marsden, J. E., \& West, M. (2001). Discrete Mechanics and Variational Integrators. Acta Numerica.\\
[5] Strogatz, S. H. (2018). Nonlinear Dynamics and Chaos. CRC Press.\\
[6] Gu, A., et al. (2021). Efficiently Modeling Long Sequences with Structured State Spaces. ICLR.\\
[7] Stürtz, J. (2026). The Runge-Kutta Paradox in Neural Manifolds. In preparation.

\end{document}
